A genome-wide MAGIC kit for recombinase-independent mosaic analysis in Drosophila

Version 1

Abstract

Abstract

Mosaic analysis has been instrumental in advancing developmental and cell biology. Most current mosaic techniques rely on exogenous site-specific recombination sequences that need to be introduced into the genome, limiting their application. Mosaic analysis by gRNA-induced crossing-over (MAGIC) was recently developed in Drosophila to eliminate this requirement by inducing somatic recombination through CRISPR/Cas9-generated DNA double-strand breaks. However, MAGIC has not been widely adopted because gRNA-markers, a required component for this technique, are not yet available for most chromosomes. Here, we present a complete, genome-wide gRNA-marker kit that incorporates optimized designs for enhanced clone induction and more effective clone labeling in both positive MAGIC (pMAGIC) and negative MAGIC (nMAGIC). With this kit, we demonstrate clonal analysis in a broad range of Drosophila tissues, including cell types that have been difficult to analyze using recombinase-based systems. Notably, MAGIC enables clonal analysis of pericentromeric genes and deficiency chromosomes and in interspecific hybrid animals, opening new avenues for gene function study, rapid gene discovery, and understanding cellular basis of speciation. This MAGIC kit complements existing systems and makes mosaic analysis accessible to address a wider range of biological questions.

Introduction

Mosaic animals containing genetically distinct populations of cells in the same organism are useful for in vivo studies of complex biological processes. For this reason, techniques that can generate genetically labeled mosaic clones have been utilized in both vertebrates and invertebrates to study tissue-specific functions of pleiotropic genes, developmental timing, cell lineages, cell proliferation, neural wiring, and many other biological phenomena [1–3]. The most popular mosaic techniques rely on site-specific recombination systems, such as FRT/Flp [4, 5] and LoxP/Cre [6, 7], to induce somatic recombination between homologous chromosomes. Such techniques require the introduction of recombination sites to specific locations in the genome and thus cannot be applied to unmodified chromosomes. To overcome this limitation, we recently developed a recombinase-independent mosaic technique called mosaic analysis by gRNA-induced crossing-over (MAGIC) [8]. In MAGIC, the CRISPR/Cas9 system generates double-strand breaks (DSBs) at a predefined genome location to induce homologous recombination in precursor cells during S/G2 phase. Subsequent chromosomal segregation during mitosis can result in clones homozygous for the chromosomal segments distal to the crossover site. Two variants of this technique in Drosophila, positive MAGIC (pMAGIC) and negative MAGIC (nMAGIC), label the resulting homozygous clones by the presence and absence of fluorescent markers, respectively [8]. Like FRT/Flp-based techniques, MAGIC enables characterization of homozygous clones of lethal mutations in otherwise heterozygous animals [8].

However, unlike FRT/Flp-based techniques, MAGIC does not require prior genetic modification of the test chromosome. Thus, it can potentially be used on any chromosome and have much wider applications. Foremost, mutations of diverse natures have been established for most Drosophila genes [9–14], and thousands of deficiency strains harbor deletions that collectively uncover 98.4% of the Drosophila genome [15]. However, most of these mutant chromosomes cannot be analyzed by traditional mosaic techniques due to the lack of FRT sites or incompatibility with the FRT/Flp system. In contrast, MAGIC can theoretically be applied to these chromosomes as they are, allowing convenient genome-wide mosaic screens. In addition, genes located more proximal to centromeres than existing FRT sites cannot be analyzed by FRT/Flp techniques. In comparison, MAGIC can potentially be used to study these genes because the crossover site in MAGIC can be flexibly defined by users. Lastly, given that MAGIC is compatible with wild-derived chromosomes [8], it may be able to generate homozygous clones of a chromosome derived from a single species in an interspecific hybrid animal, allowing the study of species-related cell-cell interactions.

Despite these potentials, MAGIC has not been widely adopted by the Drosophila community. A major barrier is the lack of gRNA-marker transgenes on most chromosomal arms, which are necessary for DSB induction and fluorescent clone labeling [8]. In addition, the existing gRNA-markers suffer from several limitations, including low frequency of clone induction, weak labeling of pMAGIC clones, and suboptimal visualization of nMAGIC clones. Because of these reasons, MAGIC has only been successfully applied to a few genes in a limited number of Drosophila tissues [8, 16].

To overcome these limitations, here we first optimized gRNA-marker designs to improve clonal induction, the brightness of pMAGIC clones, and visualization of nMAGIC clones. Then we generated pMAGIC and nMAGIC gRNA-markers for all chromosomal arms and characterized their ability to generate clones. Using this kit, we demonstrate mosaic analysis of centromere-proximal genes, deficiency chromosomes, chromosomes derived from different Drosophila species in interspecific hybrid animals. This kit allows optimal clone induction in diverse cell and tissue types and should be useful for studying a wide range of biological processes.

Results

New gRNA-marker designs improve pMAGIC and nMAGIC

MAGIC relies on a gRNA-marker dual-transgene (Figure 1A) inserted in a specific chromosomal arm to both induce and visualize clones homozygous for this arm [8]. The gRNA part of the construct expresses two gRNAs ubiquitously to target a pericentromeric location on this arm. Two gRNAs targeting two sequences that are close to each other, instead of a single gRNA, are used to increase the probability of DSBs and thus the clone frequency. The marker part in pMAGIC utilizes a ubiquitously expressed Gal80 to prevent Gal4-dependent labeling of heterozygous and homozygous cells for the gRNA-marker, while allowing labeling of homozygous cells that lose the gRNA-marker. In contrast, nMAGIC uses ubiquitously expressed BFP, which results in brighter labeling of gRNA-marker homozygous cells, intermediate labeling of gRNA-marker heterozygous cells, and lack of labeling of gRNA-marker-negative homozygous cells.

New gRNA-marker designs improve pMAGIC and nMAGIC.

(A) Original and new designs of gRNA-markers for pMAGIC and nMAGIC. (B) Comparison of clone frequency in larval sensory neurons between two gRNA designs. The number represents clones between A1 and A7 segments on one side of each larva. n = larvae number: tgFE (n=10), Qtg2.1 (n=10). (C-E) Labeling of hemocytes in whole 3rd instar larvae by pxnGal4>CD4-tdTom alone (C) or together with ubi-Gal80 (D) or tub-Gal80 (E). The panels on the right show enlarged views of the boxed regions. (F) Designs of Gal80 variants tested in pMAGIC gRNA-markers. (G) The brightness of epidermal clones labeled by pMAGIC gRNA-markers. n = image numbers: gRNA-40D2-uH (n = 32), gRNA-40D2-uDEH (n = 31), gRNA-42A4-uDEH (n = 52), gRNA-42A4-tDEH (n = 39), gRNA-42A4-tDES (n = 38). (H) The brightness of neuronal clones labeled by pMAGIC gRNA-markers. n = neuron numbers: gRNA-40D2-uH (n = 16), gRNA-40D2-uDEH (n = 16), gRNA-42A4-uDEH (n = 16), gRNA-42A4-tDEH (n = 15), gRNA-42A4-tDES (n = 16). (I) Portion of a larval wing disc containing nMAGIC clones visualized by nlsBFP. (J and J’) Portion of a wing disc containing nMAGIC clones labeled by cytosolic BFP (J) and HA staining (J’). (K) Epidermal clones on the larva body wall labeled by nlsBFP. (L) Epidermal clones visualized by cytosolic BFP. In all plots, black bar, mean; red bar, SD; AU, arbitrary unit. Student’s t-test in (B); one-way analysis of variance (ANOVA) and Tukey’s honest significant difference (HSD) test in (G) and (H). *p≤0.05, **p≤0.01, ***p≤0.001, ns, not significance. For (C-E), scale bar, 300 µm. For (I-M), scale bar, 100 µm.

We have previously developed gRNA-marker vectors for both pMAGIC and nMAGIC. However, gRNA-markers made with these vectors exhibit some limitations. First, the clone frequency can be low for certain gRNAs [8]. Second, the ubi-Gal80 in pMAGIC gRNA-markers does not completely suppress Gal4 activity in certain tissues, such as hemocytes. Third, pMAGIC clones are sometimes too dim to visualize cell morphology, such as thin dendrite and axon projections of neurons. Lastly, nMAGIC utilizes a nuclear BFP marker (nlsBFP), which does not show the cell shape and sometimes cannot mark clones effectively. Thus, to optimize MAGIC, we first sought to improve the gRNA-marker designs.

Since clone induction in MAGIC depends on gRNA-induced DNA DSBs, we asked whether a more efficient gRNA design enhances clone frequency in somatic tissues. The previous gRNA-marker vectors used a tgFE design, which contains the “F+E” gRNA scaffold and a tRNAGly before each gRNA targeting sequence (Figure 1A) [8]. However, we have shown that an improved Qtg2.1 design, which contains the gRNA2.1 scaffold and a single tRNAGln spacer between the two gRNAs (Figure 1A), is much more mutagenic than tgFE in somatic tissues [17]. We thus compared two pMAGIC gRNA-markers that are based on these two gRNA designs but target the same genomic sequences at cytological band 40D2 to assess their ability to induce clones in peripheral sensory neurons. When used with the same neuronal/epidermal precursor Cas9, zk-Cas9 [8], Qtg2.1 resulted in three times more neuronal clones as compared to tgFE (Figure 1B), confirming a positive correlation between gRNA efficiency and clone frequency.

To ensure more complete suppression of Gal4 activity by Gal80 in pMAGIC, we replaced the ubi enhancer driving Gal80 expression with an αTub84B (tub) enhancer that was used in tub-Gal80 in the MARCM system [18]. When combined with the larval hemocyte marker pxn-Gal4 UAS-CD4-tdTom [19] (Figure 1C), ubi-Gal80 did not completely suppress the labeling of hemocytes (Figure 1D). In contrast, no labeled hemocytes could be detected in the presence of tub-Gal80 (Figure 1E), suggesting that tub-Gal80 is a better marker for pMAGIC.

To improve clone brightness in pMAGIC, we sought to destabilize Gal80 and reduce its expression, reasoning that dim clones are due to prolonged Gal80 activity after clone induction. In addition to replacing the ubi enhancer with the tub enhancer, we also introduced protein and mRNA destabilization sequences (DE) [20, 21] at the 3’ end of the Gal80 coding sequence, and replaced the His2Av polyA by SV40 polyA (Figure 1F), as the latter reduces transgene expression [22]. By measuring the brightness of epidermal (Figure 1G) and neuronal (Figure 1H) clones induced by gRNA-40D2 and gRNA-42A4, we found that each of the three changes improved clone brightness.

Lastly, to visualize cell shape in nMAGIC, we replaced nlsBFP with cytosolic BFP tagged by 3X Hemagglutinin (HA), in addition to utilizing the tub enhancer. This new design allowed us to better discern clone shapes in both the wing imaginal disc and the epidermis (Figure 1J, 1J’, and 1L), in contrast to the previous ubi-nlsBFP design (Figure 1I and 1K).

Thus, by altering the designs of the gRNAs and the Gal80 and BFP markers, we created new vectors optimized for more robust applications of nMAGIC and pMAGIC.

A gRNA-marker kit is established for all four chromosomes of Drosophila

To enhance the utility of MAGIC in Drosophila, we generated complete sets of pMAGIC and nMAGIC gRNA-markers for all four chromosomes (Figure 2A). To identify suitable gRNA target sites, we analyzed the pericentromeric sequences of X, 2L, 2R, 3L, 3R, and 4, based on three criteria [8]: conserved in closely related Drosophila species to minimize the chance of single nucleotide polymorphism, (2) located away from functionally critical regions to avoid disrupting essential processes, and (3) unique within the genome to minimize off-target effects. For each MAGIC construct, we selected a pair of non-repetitive gRNA target sequences in intergenic regions to maximize the chances of DSBs. These two sequences are closely linked to minimize the risk of large deletions. Given the variable efficiency of gRNA target sequences, we selected three pairs of gRNAs targeting three chromosomal locations for each chromosomal arm and named them according to the corresponding cytoband (Table 1).

A genome-wide gRNA-marker kit suits diverse needs of clone frequency.

(A) Scheme of gRNA-marker insertion sites and target sites on Drosophila chromosomes. (B) Comparison of clone frequencies of all pMAGIC gRNA-markers in larval sensory neurons, clones are labeled using RabX4-Gal4 UAS-MApHs (for Chromosome X, II and IV) or 21-7-Gal4 UAS-MApHs (for Chromosome III). n = larvae number: X2 (n = 10), 20F2 (n = 10), 20F1(n = 10), 40D2 (n = 20), 40D4 (n = 10), 40E1 (n = 10), 41F9 (n = 20), 41F11 (n = 10), 42A4 (n = 10), 80C1 (n = 20), 80C2 (n = 14), 80F5 (n = 15), 81F (n = 10), 82A4 (n = 10), 82C3 (n = 10), 101F1a (n = 10), 101F1b (n = 10), 101F1c (n = 10). (C) Comparison of clone areas in larval wing discs labeled by nMAGIC gRNA-markers on 2R. n = wing disc number: 41F9 (n = 14), 41F11 (n = 16), 42A4 (n = 15). (D and E) Neuronal clones in the central part of the adult brain induced by pMAGIC gRNA-markers gRNA-40D2 (D) and gRNA-40E1 (E). In all plots, black bar, mean; red bar, SD. One-way ANOVA and Tukey’s HSD test. *p≤0.05, **p≤0.01, ***p≤0.001, ns, not significance. For (D) and (E), scale bar 100 µm.

To evaluate the clone-induction properties of these gRNAs, we combined the pMAGIC set with zk-Cas9 and counted the number of neuronal clones in A1-A7 larval hemi-segments (Figure 2B). As expected, the clone frequency varied from gRNA to gRNA, but we were able to identify efficient gRNAs for every chromosomal arm. We previously found that different gRNAs follow the same trend of relative efficiency in different tissues [8]. Here we additionally tested nMAGIC gRNAs for 2R in the wing disc (Figure 2C) and noticed a similar trend of clone induction to that of their pMAGIC counterparts in sensory neurons (Figure 2B), suggesting that the results in sensory neurons are transferrable to other tissues.

Certain gRNAs (e.g. gRNA-40E1) exhibited very low clone frequency. Such gRNAs can be useful for inducing sparse clones in highly packed issues such as the brain. For example, using the same ey-Cas9 [23], the highly efficient gRNA-40D2 induced too many neuronal clones in the adult brain for morphological analysis (Figure 2D), while gRNA-40E1 gave rise to few clones, whose projection patterns were much easier to analyze (Figure 2E).

Together, the pMAGIC and nMAGIC gRNA-marker lines constitute a complete kit (Table 1) for genome-wide MAGIC applications in Drosophila.

MAGIC allows clonal analysis in diverse tissues and cell types

To determine if MAGIC can be applied to diverse tissues in Drosophila, we conducted clonal analysis in the larva using gRNA-40D2(Gal80) and several tissue-specific Cas9s. With the ubiquitous vas-cas9 [24] and tub-Gal4 UAS-mCD8-GFP, we readily detected clones in the larval brain (Figure 3A), proliferating tissues like eye and leg discs (Figure 3B-C), and polyploid tissues like the fat body, gut, and trachea (Figure 3D-F). Clone induction in polyploid tissues suggests that crossing-over events occurred before the last cell division. Using zk-cas9 and R38F11-Gal4>UAS-tdTomato (tdTom), we observed frequent epidermal clones (Figure 3G). Using the glia precursor gcm-Cas9 and repo-Gal4>UAS-mCD8-GFP, we detected individual glial clones in the brain (Figure 3H). Lastly, the new pMAGIC gRNA-marker design allowed us to reliably induce hemocyte clones (Figure 3I). These results collectively demonstrate MAGIC’s efficacy in generating clones in diverse Drosophila tissues, indicating its value in studying gene function and cell lineage in various tissue contexts.

MAGIC allows clonal analysis in diverse tissues and cell types.

(A-F) pMAGIC clones induced in different tissues by vas-Cas9 gRNA-40D2(Gal80) and labeled by tub-Gal4 UAS-CD8-GFP (green). DAPI staining (white) shows all nuclei. (G) A pMAGIC epidermal clone on the larval body wall induced by zk-Cas9 gRNA-40D2(Gal80) and labeled by R38F11-Gal4 UAS-tdTom (green). Epidermal junctions are labeled by α-Catenin-GFP (white). (H) pMAGIC glia clones in the larval brain induced by gcm-Cas9 gRNA-40D2(Gal80) and labeled by repo-Gal4 UAS-CD8-GFP (green). Glial nuclei are labeled by Repo staining (white). (I) pMAGIC hemocyte clones induced by Act-Cas9 gRNA-40D2(Gal80) and labeled by pxn-Gal4 UAS-tdTom. For figure A, D-F, H, scale bar 100 µm. For figure B-C, G, scale bar 50 µm. For Figure I, scale bar 25 µm.

The neuromuscular junction (NMJ) in Drosophila larvae has been a valuable model for elucidating synaptic biology, owing to its simplicity, accessibility, and conserved features shared with mammalian excitatory synapses [25–27]. Despite its popularity, the NMJ has rarely been analyzed by the MARCM technique in the literature, likely due to the difficulty of inducing clones in motor neurons. To test the effectiveness of MAGIC in analyzing gene function at the NMJ, we selected two genes crucial for synaptic function and construction: Vesicular glutamate transporter (VGlut) [28] and bruchpilot (brp) [29, 30], null mutations of each of which exhibit recessive lethality in larvae. Combined with appropriate gRNA(Gal80) lines (42A4 for vGluT on 2R and 40D2 for brp on 2L) and tub-Gal4 UAS-CD8-GFP, zk-Cas9 induced frequent clones in type Ib boutons (Figure 4A-B”). The loss of vGluT or Brp specifically in GFP-labeled NMJs was confirmed by immunostaining. These results exemplify the value of pMAGIC for dissecting gene function in NMJ biology at the single cell level.

MAGIC facilitates clonal analysis at the NMJ.

(A-A”) pMAGIC clones of VGlut1 mutation in motor neurons at the neuromuscular junction. Clones were induced by zk-Cas9 gRNA-40D2(Gal80) and labeled by tub-Gal4 UAS-CD8-GFP., The loss of VGlut is confirmed by VGlut staining. The mutant clones are outlined in (A”). (B-B”) A pMAGIC clone of brpd09839 mutation in a motor neuron at the neuromuscular junction. Clones were induced by zk-Cas9 gRNA-42A4(Gal80) and labeled by tub-Gal4 UAS-CD8-GFP. The loss of Brp is confirmed by Brp staining. The mutant clone is outlined in (B”). In both experiments, HRP staining shows all axons. Scale bars, 10 µm.

MAGIC enables clonal analysis of pericentromeric genes, 4th chromosome-associated mutations, and in interspecific hybrid animals

Because all FRT sites in the existing FRT/Flp mosaic system are located at some distances away from centromeres, it has not been possible to study genes located between the FRT sites and the corresponding centromeres by clonal analysis. In contrast, the gRNA target site in a MAGIC experiment can be user-selected to enable clonal analysis of any given gene. To illustrate this potential, we used gRNA-41F9 to induce homozygous mutant clones of Ecdysone receptor (EcR), which is located at 42A10 and is inaccessible to all FRT sites on 2R. EcR is a transcription factor required for neuronal remodeling during metamorphosis [31]. In peripheral sensory neurons, EcR activation at the beginning of pupariation causes apoptosis of some dendritic arborization (da) neurons while triggering dendrite pruning of other da neurons [32]. As expected, a wildtype (WT) pMAGIC clone of class IV da (C4da) neuron exhibited complete dendrite pruning at 16 h after puparium formation (APF), accompanied by dendrite debris phagocytosed into epidermal cells [19] (Figure 5A). In contrast, EcR mutant da neuron clones still maintained larval dendritic arbors at 16 h APF (Figures 5B-D), instead of dying (in the case of C3da) or undergoing dendrite pruning (in the cases of C1da and C4da).

MAGIC enables clonal analysis of pericentromeric genes, 4th chromosome-associated mutations, and in interspecific hybrid animals.

(A) A WT pMAGIC class IV da neuron clone exhibiting complete dendrite pruning at 16 hours APF. (B-D) pMAGIC clones of EcRM554fs mutation in da neurons imaged at 16 hours APF, exhibiting the lack of pruning (B and D) or apoptosis (C). In (A-D), the clones were induced by zk-cas9 with gRNA-41F9(Gal80) and labeled by RabX4-Gal4 UAS-MApHS. Neuronal cell bodies are indicated by arrows. MApHS contains pHluorin and tdTom [19], but only tdTom signals are shown. The signals in epidermal cells (A) were due to engulfment of pruned dendrites by epidermal cells [19]. (E and F) WT (E) and Df(4)ED6380 (F) pMAGIC clones in C4da neurons induced by zk-cas9 gRNA-101Fc(Gal80) and labeled by RabX4-Gal4 UAS-MApHS. Only tdTom signals are shown. (G) Normalized dendrite length of WT clones and deficiency clones. Black bar, mean; red bar, SD. Student’s t-test. ***p≤0.001. (H) Scheme for interspecific crosses between D. melanogaster (D.m) and D. simulans (D.s). (I and J) Wing discs from male (I) and female (J) progeny carrying clones. Scale bars, 50 µm.

Mosaic analysis of genes located on the 4th chromosome had not been possible until recent introduction of FRT sites onto this chromosome by CRISPR-mediated knock-in [33]. Despite these advances, existing mutations on FRT-lacking 4th chromosomes still cannot be analyzed by the FRT/Flp system. A valuable gene-disruption resource in Drosophila is the deficiency library consisting of strains harboring chromosomal deletions. MAGIC can potentially be used in conjunction with the deficiency library to screen for genes important in a particular biological process. To test the potential of MAGIC for analyzing mutations on the 4th chromosome and for gene discovery with deficiencies, we generated pMAGIC C4da clones of Df(4)ED6380, a deficiency that deletes a segment between cytological bands 102B7 and 102D5. These clones show dramatically reduced dendrites compared to the WT controls (Figure 4E-4G), indicating the existence of genes important for dendrite growth in this region.

When examining nMAGIC gRNA-markers for the 4th chromosome, we noticed uneven expression of tub-3xHA-BFP in epidermal cells and some imaginal tissues, exemplified by cells lacking detectable BFP signal (Figures S1A,S1B). This variegated expression is likely due to transgenes residing in heterochromatin and is common for transgenes located on the 4th chromosome [34]. While this uneven expression limits the usefulness of our gRNA-markers in the corresponding epithelial tissues, Gal80 in pMAGIC gRNA-markers efficiently suppressed Gal4 activities in most neurons (Figure S1C), conforming their effectiveness in neuronal MAGIC analysis.

Lastly, we wondered if the MAGIC system can generate clones in interspecific hybrid animals derived from D. melanogaster and D. simulans parents, given that the two species show a large degree of synteny [35]. The clones in these animals would contain homozygous chromosomal arms derived from a single species. To test this idea, we crossed D. melanogaster females containing gRNA-42A4(BFP); hh-cas9 to D. simulans males carrying a loss of function mutation of Lethal hybrid rescue mutation (Lhr), which ensures the viability of the hybrid male progeny [36]. Indeed, we observed twin spots consisting of dark clones, containing only D. simulans 2R, and brighter clones, containing only D. melanogaster 2R, in wing discs of both female and male progeny (Figure 5G and 5H), demonstrating the feasibility of studying species-specific alleles in cell-cell interactions in interspecific hybrids.

Transgenic markers on the 4th chromosome show uneven expression.

(A-B) Representative epidermal (A) and wing disc (B) images showing uneven expression of gRNA-101F1c(BFP) inserted at attP102D on the 4th chromosome. Yellow arrowheads point to cells lacking BFP expression. Scale bars, 50 µm. (C) Frequency of labeled neurons by indicated gRNA(Gal80) in the absence and presence of Cas9. The zk-Cas9 dataset is the same as that for chromosome 4 in Figure 2B. Black bar, mean; red bar, SD. One-way ANOVA and HSD test. ***p≤0.001.

Discussion

Conceptually, MAGIC is a simpler and more convenient mosaic technique compared to traditional recombinase-dependent methods. However, the lack of gRNA-marker transgenes has prevented its wide application in Drosophila. In this study, we present a complete gRNA-marker kit that enables genome-wide MAGIC in Drosophila, removing this bottleneck. The new gRNA-markers incorporate optimized designs that improve clone frequency and labeling in both pMAGIC and nMAGIC. We further demonstrate the compatibility of MAGIC with broad tissues and cell types. More importantly, MAGIC enables mosaic analyses that could not be accomplished by existing FRT/Flp systems, such as those of pericentromeric genes, deficiency chromosomes, and interspecific homologous chromosomes. Thus, this MAGIC kit provides Drosophila researchers with greater flexibility for conducting mosaic analyses of diverse purposes.

To implement MAGIC, one needs to first choose a proper Cas9 that is expressed in the precursor cells of the targeted cell population. We show that ubiquitously expressed Cas9 lines, such as vas-Cas9 [24] and Act5C-Cas9 [37], are sufficient to induce clones in broad tissues. Alternatively, heat shock (HS)-induced Cas9 [38] can offer temporal control of clone induction in most tissues, akin to the HS-Flp commonly used in FRT-based mosaic analysis. The third option is a Cas9 driven by a tissue-specific enhancer specifically in the precursor cells of the target tissues. Examples shown in this study include ey-Cas9 expressed in many neuronal lineages [23], gcm-Cas9 expressed in glial progenitor cells [39], ZK-Cas9 expressed in precursor cells of epidermal cells, motor neurons, and somatosensory neurons [17], and hh-Cas9 expressed in imaginal tissues [40]. These Cas9 lines have the advantages of being more specific and efficient, and more convenient to use than ubiquitous or inducible ones. Multiple strategies, including enhancer fusion [37, 40], Gal4-to-Cas9 conversion [17, 41], and insertion of Cas9 in specific gene loci [39], have been developed to ease the generation of such tissue-specific Cas9 lines. As an ongoing effort, we have been converting Gal4 lines known to be expressed in progenitor cells into Cas9 (available at https://bdsc.indiana.edu/stocks/genome_editing/crispr_cas9.html), in the hope of making MAGIC accessible to more Drosophila tissues.

The second component of MAGIC is a gRNA-marker line that resides on the appropriate chromosome arm and targets Cas9 to cut a pericentromeric site on the same arm. In the gRNA-marker kit, we generated three lines targeting different sites for each chromosome arm. These lines exhibit a broad range of clone frequencies in larval sensory neurons. As different gRNA-markers maintain similar relative efficiencies of clone induction across tissues [8], one can choose a gRNA-marker more appropriate for their applications based on the target gene location and the desired clone frequency. Notably, a higher clone frequency may not always be beneficial, such as when studying the projection patterns of individual neurons in the brain. A low efficiency gRNA-marker in this kit may be more desirable in such applications.

Besides commonly conducted mosaic analysis in Drosophila, MAGIC also enables many novel analyses that are difficult or impossible to accomplish with traditional systems. One example is to analyze interactions among species-specific cells in an interspecific hybrid animal for understanding the cell biological basis of hybrid incompatibility. Interspecific crosses between WT D. melanogaster and WT D. simulans result in sex-specific lethality of F1 progeny [36]. The cellular basis of this lethality is poorly understood. With MAGIC, one may generate mixed cell populations that are homozygous or heterozygous for species-specific alleles in the same hybrid embryo or larva, in which the relative fitness of the three cell populations can be compared. Similarly, MAGIC could complement genome-wide association studies (GWAS) using wild-derived isogenic strains and provide much more mechanistic detail. Such strains as those in the Drosophila Genetic Reference Panel (DGRP) [42] have been very useful for discovering loci that are responsible for phenotypic variations. MAGIC can be combined with these strains to analyze the effects of homozygosity of specific variants at the cell biology level [8]. In addition, MAGIC can be combined with the Drosophila deficiency kit for rapid genome-wide mosaic screens. A deficiency exhibiting a desired phenotype in such screens can be further dissected with smaller deficiencies within the deleted region or existing mutations in candidate genes. In this way, the convenience of MAGIC could significantly accelerate phenotype-based gene discovery. Lastly, because the crossover site in MAGIC can be user-defined, one can easily generate gRNA-markers that induce somatic recombination at specific genome locations. Such flexibility could be useful for mapping undefined genetic loci that are responsible for certain phenotypes, especially in non-traditional model organisms.

Materials and methods

Fly Stocks and Husbandry

See the Key Resource Table (Table S1) for details of fly stocks used in this study. Most fly lines were either generated in the Han lab or obtained from the Bloomington Drosophila Stock Center. lethal hybrid rescue (lhr) mutant D. Simulans was a gift from Dr. Dan Barbash. All flies were grown on standard yeast-glucose medium, in a 12:12 light/dark cycle, at 25°C unless otherwise noted. Virgin males and females for mating experiments were aged for 3-5 days.

To generate and label pMAGIC clones in larval peripheral sensory neurons, we used either RabX4-Gal4 UAS-MApHS (for gRNA-markers on chromosomes X, II, and IV) or 21-7-Gal4 UAS-MApHS (for gRNA-markers on chromosome III) combined with zk-cas9. To count peripheral sensory neuronal clones on the larval body wall, third instar larvae were mounted laterally on slides and then counted in segment A1-A7 under a Nikon SMZ18 stereomicroscope. pMAGIC clones were induced and labeled by RabX4-Gal4 UAS-MApHS combined with ey-cas9 in the fly adult brain, by tub-Gal4 UAS-mCD8-GFP combined with vas-cas9 in larval brains, imaginal discs, fat bodies, guts, and trachea, by repo-Gal4 UAS-mCD8-GFP combined with gcm-cas9 in glia, by pxn-Gal4 UAS-tdTom combined with Act-cas9 in hemocytes, by tub-Gal4 UAS-mCD8-GFP combined with zk-cas9 in larval motor neurons, by R38F11-Gal4 UAS-tdTom combined with zk-cas9 in the larval epidermis. To induce nMAGIC clones in wing imaginal discs of interspecific hybrid animals, we use gRNA-42A4(BFP); hh-cas9 virgin females of D. melanogaster to cross with Lhr1 [43] D. simulans males.

Molecular Cloning

MAGIC gRNA cloning vectors: pMAGIC gRNA-marker vectors constructed in this study include pAC-U63-gRNA2.1-ubiGal80(DE)-His2Av, pAC-U63-gRNA2.1-tubGal80(DE)-His2Av and pAC-U63-gRNA2.1-tubGal80(DE)-SV40. To make pAC-U63-gRNA2.1-ubiGal80(DE)-His2Av, a fragment containing gRNA2.1 scaffold (gRNA2.1) and U6 3’ flanking sequence (U63fl) was first used to replace gRNA2.1-QtRNA-Gal80(TS)-gRNA2.1-U63fl in pAC-U63-QtgRNA2.1-8R (Addgene 170514) [17]. Then a destabilization sequence (DE) containing a GS linker, amino acids (AAs) 422-461 of mouse ornithine decarboxylase [21], 2X RNA destabilizing nonamer (TTATTTATTgatccTTATTTATT) [21] was added to the C-terminus of Gal80 in frame. To make pAC-U63-gRNA2.1-tubGal80(DE)-His2Av, a 2.6 kb tub enhancer was amplified from pENTR221-tubP [16] by PCR and used to replace the ubi enhancer in pAC-U63-gRNA2.1-ubiGal80(DE)-His2Av. To make pAC-U63-gRNA2.1-tubGal80(DE)-SV40, a SV40 polyA sequence was amplified from pAPIC-PHCS [22] and used to replace the His2Av polyA in pAC-U63-gRNA2.1-ubiGal80(DE)-His2Av. To make the nMAGIC gRNA-marker vector pAC-U63-gRNA2.1-tubBFP(HA), the gRNA2.1-U63fl fragment was first used to replace gRNA2.1-QtRNA-BFP(TS)-gRNA2.1-U63fl in pAC-U63-QtgRNA2.1-BR (Addgene 170513) [17]. The tub enhancer was then used to replace the ubi enhancer. Lastly, a synthetic DNA fragment (GenScript) encoding 3X HA was used to replace the nuclear localization signal at the N-terminus of mTagBFP. Cloning was carried out by ligation with T4 ligase or NEBuilder DNA Assembly reactions (New England Biolabs Inc.).

MAGIC gRNA expression vectors: 33 gRNA expression vectors were constructed with the corresponding gRNA cloning vectors as listed in Table 1 according to published protocols [17]. Briefly, for each expression vector, two primers containing appropriate gRNA target sequences (Table S2) were used to amplify a fragment consisting of 3’ end of U6:3 promotor, the first target sequence (TS1), gRNA2.1, tRNAQ, the second target sequence (TS2), and the beginning sequence of gRNA2.1 with pAC-U63-QtgRNA2.1-BR as the PCR template. The PCR product was then assembled with SapI-digested gRNA cloning vectors using NEBuilder DNA Assembly.

Injections were carried out by Rainbow Transgenic Flies (Camarillo, CA 93012 USA) or Genetivision (Stafford, TX 77477) to transform flies through φC31 integrase-mediated integration into attP docker sites. The 3xP3-RFP selection marker in gRNA-markers inserted at the attP[ZH-102D] site were removed by crossing to Cre.

Live Imaging

Live imaging of larval epidermal cells, sensory neurons, and hemocytes was performed as previously described [44]. Animals were collected at 96 (for late third larvae) or 120 hours AEL (for wandering third instar larvae) hours after egg-laying (AEL) and mounted in glycerol on a slide with vacuum grease as a spacer. Animals were imaged using a Leica SP8 confocal microscope with a 40X NA1.3 oil objective, pinhole size 2 airy units, and a z-step size of 1 µm. For epidermis, images were taken at the dorsal midline of A2 and A3 segments. For dendritic arborization neurons, images were taken from A1 to A7 hemi-segments.

For imaging sensory neurons in pupae, newly formed pupae were collected and incubated at 25 °C. After 16 hours, the pupal cases were carefully removed, and the pupae were mounted dorsal side up on slides with halocarbon oil beneath a coverslip. Double-sided tape was used as a spacer. The mounted pupae were imaged using a Leica SP8 confocal microscope with a 40X NA1.3 oil objective.

Imaging of wing discs and other larval tissues

Larval dissections were performed as described previously [40]. Briefly, wandering third instar larvae were dissected in a small petri dish filled with cold phosphate-buffered saline (PBS). The anterior half of the larva was inverted. To prepare imaginal discs, trachea and gut were removed. Samples were then transferred to 4% formaldehyde in PBS and fixed for 20 minutes at room temperature. After washing with PBS, the tissues were stained in DAPI (1:1000) in 0.2% PBST (PBS with 0.2% Triton X-100) for 5 minutes. The tissues were washed again in PBST and mounted in SlowFade Diamond Antifade Mountant (Thermo Fisher Scientific) on a glass slide.

A coverslip was lightly pressed on top. Imaginal discs were imaged using a Leica SP8 confocal microscope with a 20X NA0.8 oil objective.

Adult brain imaging

Flies were aged for 1 day after eclosion. Brains were dissected in PBS at room temperature and then fixed in 4% paraformaldehyde in PBS with constant circular rotation for 20 minutes at room temperature. The brains were subsequently washed in 0.2% PBST and mounted on a glass slide under a glass coverslip. Vacuum grease was used as a spacer between the coverslip and the slide. Brains were imaged using a Leica SP8 confocal microscope with a 40X NA1.3 oil objective.

Larval fillet preparation

Larval fillet dissection was performed on a petri dish half-filled with PMDS gel. Wandering third instar larvae were pinned on the dish in PBS dorsal-side up and then dissected to expand the body wall. PBS was then removed and 4% formaldehyde in PBS was added to fix larvae for 15 minutes at room temperature. For VGlut staining, the fillets were fixed in Bouin’s solution for 5 minutes at room temperature. Fillets were rinsed and then washed at room temperature in PBS for 20 minutes or until the yellow color from Bouin’s solution faded. After immunostaining, the head and tail of fillets were removed, and the remaining fillets were placed in SlowFade Diamond Antifade Mountant on a glass slide. A coverslip was lightly pressed on top. Larval fillets were imaged using a Leica SP8 confocal microscope with a 40X NA1.3 oil objective.

Immunohistochemistry

Larval brains and larval fillets were rinsed and washed at room temperature in 0.2% PBST after fixation. The samples were then blocked in PBST with 5% normal donkey serum (NDS) for 1 hour before incubating with appropriate primary antibodies in the blocking solution. Brains were stained for 2 hours at room temperature, and fillets were stained overnight at 4°C. After additional rinsing and washing, the samples were incubated with secondary antibodies for 2 hours at room temperature. The samples were then rinsed and washed again before mounting and imaging. Primary antibodies used in this study are mouse anti-Repo antibody 8D12 (1:50 dilution), mouse anti-Brp antibody nc82 (1:100 dilution), rabbit anti-VGlut (1:200 dilution) [39], mouse anti-HA antibody (12CA5, 1:100), and goat anti-HRP conjugated with Cy3 (1:200). Secondary antibodies include donkey anti-mouse antibody conjugated with Cy5 (1:400) and donkey anti-rabbit antibody conjugated with Cy5 (1:400).

Image Analysis and Quantification

ImageJ analyses were conducted in Fiji/ImageJ. To compare brightness of clones in sensory neurons and epidermal cells, clones were detected based on thresholds to generate masks. The clone brightness within the masks was then measured as mean pixel intensity. To quantify clones in wing discs, 2-3 slices of optical sections in the middle of the disc were projected into a 2-dimensional (2D) image. Clones were detected based on a fixed threshold to generate masks, and the total area of clones in the wing pouch of each disc was measured.

The tracing and measurement of da neuron dendrites were done as previously described in detail [44]. Briefly, dendrites were segmented using local thresholding. The segments were then converted into single-pixel-width skeletons. The total length of skeletons was calculated based on pixel distance. Normalized dendrite length was calculated as dendritic length (μm)/segment width (μm).

Statistical Analysis

One-way analysis of variance (ANOVA) with Tukey’s honest significant difference (HSD) test was used when dependent variable was normally distributed and there was approximately equal variance across groups. Student’s t-test was used when two groups are compared. For additional information on the number of samples, see figure legends. R studio was used for all statistical analyses.

eLife Assessment

The study showcases a significant and important enhancement of the MAGIC transgenesis method, by extending it genome-wide to all chromosomes. The authors convincingly demonstrate that the MAGIC mosaic clones can be generated for genes from all, including the 4th chromosome. With this toolkit extension, the method is now most likely set to strongly rival the classical FRT/Flp recombination system for gene manipulation in flies.

Reviewer #1 (Public review):

Summary:

In this manuscript, Shen et al. have improved upon the mitotic clone analysis tool MAGIC that their lab previously developed. MAGIC uses CRISPR/Cas9-mediated double-stranded breaks to induce mitotic recombination. The authors have replaced the sgRNA scaffold with a more effective scaffold to increase clone frequency. They also introduced modifications to positive and negative clonal markers to improve signal-to-noise and mark the cytoplasm of the cells instead of the nuclei. The changes result in increase in clonal frequencies and marker brightness. The authors also generated the MAGIC transgenics to target all chromosome arms and tested the clone induction efficacy.

Strengths:

MAGIC is a mitotic clone generation tool that works without prior recombination to special chromosomes (e.g., FRT). It can also generate mutant clones for genes for which the existing FRT lines could not be used (e.g., the genes that are between the FRT transgene and the centromere).

This manuscript does a thorough job in describing the method and provides compelling data that support improvement over the existing method.

Weaknesses:

It would be beneficial to have a greater variety of clonal markers for nMAGIC. Currently, the only marker is BFP, which may clash with other genetic tools (e.g., some FRET probes) depending on the application. It would be nice to have far-red clonal markers.

Reviewer #2 (Public review):

Summary:

In this study, the authors present the latest improvement of their previously published methods, pMAGIC and nMAGIC, which can be used to engineer mosaic gene expression in wild-type animals and in a tissue-specific manner. They address the main limitation of MAGIC, the lack of gRNA-marker transgenes, which has hampered the broader adoption of MAGIC in the fly community. To do so, they create an entire toolkit of gRNA markers for every Drosophila chromosome and test them across a range of different tissues and in the context of making Drosophila species hybrid mosaic animals. The study provides a significant and broadly useful improvement compared to earlier versions, as it broadens the use-cases for transgenic manipulation with MAGIC to virtually any subfield of Drosophila cell biology.

Strengths:

Major improvements to MAGIC were made in terms of clone induction efficiency and usability across the Drosophila model system, including wild-type genotypes and the use in non-melanogaster species.

Notably, mosaic mutants can now be created for genes residing on the 4th chromosome, which is exciting and possibly long-awaited by 4th chromosome gene enthusiasts.

Selection of the standard set of gRNA markers was done thoughtfully, using non-repetitive conserved and unique sequences.

The authors demonstrate that MAGIC can be used easily in the context of interspecific hybrids. I believe this is a great advancement for the Drosophila community, especially for evolutionary biologists, because this may allow for easy access to mechanistic, tissue-specific insight into the process of a range of hybrid incompatibilities, an important speciation process that is normally difficult to study at the level of molecular and cell biology.

In the same way, because it is not limited to usage in any particular genetic background, genome-wide MAGIC can be potentially used in wild-type genotypes relatively easily. This is exciting, especially because natural genetic diversity is rarely investigated more mechanistically and at the scale/resolution of cells or specific tissues. Now, one can ask how a particular naturally occurring allele influences cell physiology compared to another (control) while keeping the global physiological context of the particular genetic background largely intact.

Weaknesses:

It is not entirely clear how functionally non-critical regions were evaluated, besides that they are selected based on conservation of sequence between species. It may be useful to directly test the difference in viability or other functionally relevant phenotype for flies carrying different markers. Similarly, the frequency of off-targets could be investigated or documented in a bit more detail, especially if one of the major use-cases is meant for naturally derived, diverse genetic backgrounds. It is, at the moment, unclear how consistently the clones are induced for each new gRNA marker across different WT genetic backgrounds, for example, a set of DGRP genotypes, which could be highly useful information for future users.

Reviewer #3 (Public review):

Summary:

In the manuscript by Shen, Yeung, and colleagues, the authors generate an improved and expanded Mosaic analysis by gRNA-induced crossing-over (MAGIC) toolkit for use in making mosaic clones in Drosophila. This is a clever method by which mitotic clones can be induced in dividing cells by using CRISPR/Cas9 to generate double-strand breaks at specific locations that induce crossing over at those locations. This is conceptually similar to previous mosaic methods in flies that utilized FRT sites that had been inserted near centromeres along with heat-shock inducible FLPase. The advantage of the MAGIC system is that it can be used along with chromosomes lacking FRT sites already introduced, such as those found in many deficiency collections or in EMS mutant lines. It may also be simpler to implement than FRT-based mosaic systems. There are two flavors of the MAGIC system: nMAGIC and pMAGIC. In nMAGIC, the main constituents are a transgene insertion that contains gRNAs that target DNA near the centromere, along with a fluorescent marker. In pMAGIC, the main constituents are a transgenic insertion that contains gRNAs that target DNA near the centromere, along with ubiquitous expression of GAL80. As such, nMAGIC can be used to generate clones that are not labelled, whereas pMAGIC (along with a GAL4 line and UAS-marker) can be used much like MARCM to positively label a clone of cells. This manuscript introduces MAGIC transgenic reagents that allow all 4 chromosomes to be targeted. They demonstrate its use in a variety of tissues, including with mutants not compatible with current FLP/FRT methods, and also show it works well in tissues that prove challenging for FLP/FRT mosaic analyses (such as motor neurons). They further demonstrate that it can be used to generate mosaic clones in non-melanogaster hybrid tissues. Overall, this work represents a valuable improvement to the MAGIC method that should promote even more widespread adoption of this powerful genetic technique.

Strengths:

(1) Improves the design of the gRNA-marker by updating the gRNA backbone and also the markers used. GAL80 now includes a DE region that reduces the perdurance of the protein and thus better labeling of pMAGIC clones. The data presented to demonstrate these improvements is rigorous and of high quality.

(2) Introduces a toolkit that now covers all chromosome arms in Drosophila. In addition, the efficiency of 3 target different sites is characterized for each chromosome arm (e.g., 3 different gRNA-Marker combinations), which demonstrate differences in efficiency. This could be useful to titrate how many clones an experimenter might want (e.g., lower efficiency combinations might prove advantageous).

(3) The manuscript is well written and easy to follow. The authors achieved their aims of creating and demonstrating MAGIC reagents suitable for mosaic analysis of any Drosophila chromosome arm.

(4) The MAGIC method is a valuable addition to the Drosophila genetics toolkit, and the new reagents described in this manuscript should allow it to become more widely adopted.

Weaknesses:

(1) The MAGIC method might not be well known to most readers, and the manuscript could have benefited from schematics introducing the technique.

(2) Traditional mosaic analyses using the FLP/FRT system have strongly utilized heat-shock FLPase for inducible temporal control over mitotic clones, as well as a way to titrate how many clones are induced (e.g., shorter heat shocks will induce fewer clones). This has proven highly valuable, especially for developmental studies. A heat-shock Cas9 is available, and it would have been beneficial to determine the efficiency of inducing MAGIC clones using this Cas9 source.

Author response:

The following is the authors’ response to the original reviews.

We greatly appreciate the reviewers’ constructive comments and have followed their recommendations to improve our manuscript. These improvements include additional experiments, new analyses, and a rewriting of the text. We believe these changes significantly improved the paper and hope the editor and the reviewers agree. The following is a summary of the major changes made and our point-by-point response to reviewers’ comments.

Summary of major changes:

(1) Expanded labeling options: We generated a new nMAGIC vector containing miRFP680 as an infrared fluorescent protein (IFP) marker. We used gRNA-40D2(IFP) to demonstrate clones labeled by this marker in the wing imaginal disc (Figure 1M). This vector is available via Addgene for the generation of new gRNA-markers with our recommended or customer-designed gRNA target sequences.

(2) Validated Gal80 potency: We provide new data in Figure 1E demonstrating complete suppression of pxn-Gal4>CD4-tdTom by tub-GAL80-DE-SV40. The exact transgenes used in the comparisons are clarified in the figure and figure legend.

(3) Verified clone fitness: We compared the sizes of nMAGIC twin spots in wing discs and found no intrinsic growth or viability bias between marker/marker and WT/WT clones (Figure 1O).

(4) Methodological Schematics: We added supplemental figures to Figure 1 to illustrate the principle of MAGIC, the difference between pMAGIC and nMAGIC, and an example of pMAGIC crossing scheme.

(5) Inducible induction: We provide new data (Figure 3J-K’) showing the induction of sparse neuronal clones in the adult brain by heat shock (hs)-Cas9.

(6) We revised texts to incorporate all other recommendations suggested by the reviewers. We also made other small changes to the manuscript to improve its readability.

Public Reviews:

Reviewer #1 (Public review):

Summary:

In this manuscript, Shen et al. have improved upon the mitotic clone analysis tool MAGIC that their lab previously developed. MAGIC uses CRISPR/Cas9-mediated double-stranded breaks to induce mitotic recombination. The authors have replaced the sgRNA scaffold with a more effective scaffold to increase clone frequency. They also introduced modifications to positive and negative clonal markers to improve signal-to-noise and mark the cytoplasm of the cells instead of the nuclei. The changes result in increase in clonal frequencies and marker brightness. The authors also generated the MAGIC transgenics to target all chromosome arms and tested the clone induction efficacy.

Strengths:

MAGIC is a mitotic clone generation tool that works without prior recombination to special chromosomes (e.g., FRT). It can also generate mutant clones for genes for which the existing FRT lines could not be used (e.g., the genes that are between the FRT transgene and the centromere).

This manuscript does a thorough job in describing the method and provides compelling data that support improvement over the existing method.

Weaknesses:

It would be beneficial to have a greater variety of clonal markers for nMAGIC. Currently, the only marker is BFP, which may clash with other genetic tools (e.g., some FRET probes) depending on the application. It would be nice to have far-red clonal markers.

We thank the reviewer for the positive comments about our study. We agree with the reviewer that adding a far-red option for nMAGIC increases the flexibility of this method. We replaced the BFP coding sequence in the nMAGIC cloning vector pAC-U63-QtgRNA2.1-tubBFP(HA) with that of miRFP680-T2A-HO1. We then used the resulting cloning vector to make a gRNA-40D2(IFP) transgene and tested it in the wing disc. Result showing clones in the wing disc are now in Figure 1M. The new cloning vector, along with others reported in our study, are available from Addgene.

Reviewer #2 (Public review):

Summary:

In this study, the authors present the latest improvement of their previously published methods, pMAGIC and nMAGIC, which can be used to engineer mosaic gene expression in wild-type animals and in a tissue-specific manner. They address the main limitation of MAGIC, the lack of gRNA-marker transgenes, which has hampered the broader adoption of MAGIC in the fly community. To do so, they create an entire toolkit of gRNA markers for every Drosophila chromosome and test them across a range of different tissues and in the context of making Drosophila species hybrid mosaic animals. The study provides a significant and broadly useful improvement compared to earlier versions, as it broadens the use-cases for transgenic manipulation with MAGIC to virtually any subfield of Drosophila cell biology.

Strengths:

Major improvements to MAGIC were made in terms of clone induction efficiency and usability across the Drosophila model system, including wild-type genotypes and the use in non-melanogaster species.

Notably, mosaic mutants can now be created for genes residing on the 4th chromosome, which is exciting and possibly long-awaited by 4th chromosome gene enthusiasts.

Selection of the standard set of gRNA markers was done thoughtfully, using non-repetitive conserved and unique sequences.

The authors demonstrate that MAGIC can be used easily in the context of interspecific hybrids. I believe this is a great advancement for the Drosophila community, especially for evolutionary biologists, because this may allow for easy access to mechanistic, tissue-specific insight into the process of a range of hybrid incompatibilities, an important speciation process that is normally difficult to study at the level of molecular and cell biology.

In the same way, because it is not limited to usage in any particular genetic background, genome-wide MAGIC can be potentially used in wild-type genotypes relatively easily. This is exciting, especially because natural genetic diversity is rarely investigated more mechanistically and at the scale/resolution of cells or specific tissues. Now, one can ask how a particular naturally occurring allele influences cell physiology compared to another (control) while keeping the global physiological context of the particular genetic background largely intact.

Weaknesses:

It is not entirely clear how functionally non-critical regions were evaluated, besides that they are selected based on conservation of sequence between species. It may be useful to directly test the difference in viability or other functionally relevant phenotype for flies carrying different markers. Similarly, the frequency of off-targets could be investigated or documented in a bit more detail, especially if one of the major use-cases is meant for naturally derived, diverse genetic backgrounds. It is, at the moment, unclear how consistently the clones are induced for each new gRNA marker across different WT genetic backgrounds, for example, a set of DGRP genotypes, which could be highly useful information for future users.

We thank the reviewer for the positive comments about our study. The reviewer raises an excellent point regarding the consistency of clone induction and potential background effects in diverse genetic backgrounds. As a standard step in building the MAGIC kit, we tested all gRNA-marker transgenes with the Cas9-LEThAL assay (Poe et al., Genetics, 2019), in which the gRNA-marker transgene was crossed to lig4 Act5C-Cas9 homozygotes. All crosses led to viable and apparently healthy female progeny, suggesting that ubiquitously mutating the chosen gRNA targeting sites does not cause obvious defects.

For standard mutant analysis, we recommend researchers to use a well-characterized wildtype chromosome as a negative control. For studies utilizing diverse wildtype backgrounds where a standard control chromosome is inapplicable (e.g., DGRP screens), we recommend an internal validation strategy: researchers should confirm their key phenotypic findings by inducing clones with a second, independent gRNA-marker located on the same chromosomal arm (e.g., comparing clones induced by gRNA-40D2 vs. gRNA-40D4 ). This ensures that any observed phenotypes or variations in clone induction are linked to the selected genetic background rather than an off-target artifact or target-site specific effect.

We admit that the above approach may not resolve concerns about off-targets. Performing deep sequencing to map empirical off-targets for all 34 gRNA pairs across multiple genetic backgrounds is experimentally prohibitive for a toolkit resource. However, our in silico selection pipeline strictly required target sequences to be unique within the D. melanogaster genome to mathematically minimize off-target probability. In addition, our requirement that target sequences be conserved in closely related Drosophila species acts as a stringent filter against intraspecies variation. Sequences conserved across species are subject to purifying selection, substantially reducing the likelihood that SNPs within the DGRP lines will disrupt the PAM or seed sequences required for Cas9 induction.

Reviewer #3 (Public review):

Summary:

In the manuscript by Shen, Yeung, and colleagues, the authors generate an improved and expanded Mosaic analysis by gRNA-induced crossing-over (MAGIC) toolkit for use in making mosaic clones in Drosophila. This is a clever method by which mitotic clones can be induced in dividing cells by using CRISPR/Cas9 to generate double-strand breaks at specific locations that induce crossing over at those locations. This is conceptually similar to previous mosaic methods in flies that utilized FRT sites that had been inserted near centromeres along with heat-shock inducible FLPase. The advantage of the MAGIC system is that it can be used along with chromosomes lacking FRT sites already introduced, such as those found in many deficiency collections or in EMS mutant lines. It may also be simpler to implement than FRT-based mosaic systems. There are two flavors of the MAGIC system: nMAGIC and pMAGIC. In nMAGIC, the main constituents are a transgene insertion that contains gRNAs that target DNA near the centromere, along with a fluorescent marker. In pMAGIC, the main constituents are a transgenic insertion that contains gRNAs that target DNA near the centromere, along with ubiquitous expression of GAL80. As such, nMAGIC can be used to generate clones that are not labelled, whereas pMAGIC (along with a GAL4 line and UAS-marker) can be used much like MARCM to positively label a clone of cells. This manuscript introduces MAGIC transgenic reagents that allow all 4 chromosomes to be targeted. They demonstrate its use in a variety of tissues, including with mutants not compatible with current FLP/FRT methods, and also show it works well in tissues that prove challenging for FLP/FRT mosaic analyses (such as motor neurons). They further demonstrate that it can be used to generate mosaic clones in non-melanogaster hybrid tissues. Overall, this work represents a valuable improvement to the MAGIC method that should promote even more widespread adoption of this powerful genetic technique.

Strengths:

(1) Improves the design of the gRNA-marker by updating the gRNA backbone and also the markers used. GAL80 now includes a DE region that reduces the perdurance of the protein and thus better labeling of pMAGIC clones. The data presented to demonstrate these improvements is rigorous and of high quality.

(2) Introduces a toolkit that now covers all chromosome arms in Drosophila. In addition, the efficiency of 3 target different sites is characterized for each chromosome arm (e.g., 3 different gRNA-Marker combinations), which demonstrate differences in efficiency. This could be useful to titrate how many clones an experimenter might want (e.g., lower efficiency combinations might prove advantageous).

(3) The manuscript is well written and easy to follow. The authors achieved their aims of creating and demonstrating MAGIC reagents suitable for mosaic analysis of any Drosophila chromosome arm.

(4) The MAGIC method is a valuable addition to the Drosophila genetics toolkit, and the new reagents described in this manuscript should allow it to become more widely adopted.

Weaknesses:

(1) The MAGIC method might not be well known to most readers, and the manuscript could have benefited from schematics introducing the technique.

We thank the reviewer for the positive evaluation of our study and for making this kind suggestion. We have added diagrams that explain the principle of MAGIC and the difference between pMAGIC and nMAGIC in Figure 1 - Figure Supplement 1.

(2) Traditional mosaic analyses using the FLP/FRT system have strongly utilized heat-shock FLPase for inducible temporal control over mitotic clones, as well as a way to titrate how many clones are induced (e.g., shorter heat shocks will induce fewer clones). This has proven highly valuable, especially for developmental studies. A heat-shock Cas9 is available, and it would have been beneficial to determine the efficiency of inducing MAGIC clones using this Cas9 source.

We thank the reviewer for suggesting this experiment. We agree that demonstrating inducible clone induction in the adult brain is an effective way for people to compare MAGIC with the MARCM method they are probably more familiar with. We used a heat shock Cas9 developed by the Tzumin Lee group (Chen et al., Development, 2020) to experiment with clone induction, and the results are shown in the new Figure 3 (K and J). We show that, with a pan-neuronal Gal4, heat shock during the wandering 3rd instar larval stage induced more clones than during the pupal stage, and the later heat shock readily produced sparsely labeled neurons whose single-cell morphology can be easily visualized.

Recommendations for the authors:

Reviewing Editor Comments:

The following are some consolidated review remarks after discussions amongst all three reviewers:

The reviewers feel the evidence level could be raised from 'convincing' to 'compelling' if the following key (and partially shared) suggestions by the reviewers are followed adequately:

(1) Expand labeling options for nMAGIC, which is currently just a BFP marker. This would increase the utility of the method. A far-red marker would be very helpful. Could the authors just do this for one chromosome arm and make the reagent available for others to generate other chromosome arms?

We agree with the editor and reviewers that adding a far-red option for nMAGIC increases the flexibility of this method. We replaced the BFP coding sequence in the nMAGIC cloning vector pAC-U63-QtgRNA2.1-tubBFP(HA) with that of miRFP680-T2A-HO1. We then used the resulting cloning vector to make a gRNA-40D2(IFP) transgene and tested it in the wing disc. Result showing clones in the wing disc are now in Figure 1M. The new cloning vector, along with others reported in our study, will be available from Addgene.

(2) Verify that destabilized GAL80 is potent enough to suppress GAL4. Repeat Figure 1C-E with tub-GAL80-DE-SV40.

We replaced the experiment using gRNA-42A4-tDES, which successfully achieved complete suppression of pxn>CD4-tdTom (Figure 1E).

(3) Concern about the health of the induced mitotic clones. This is an important consideration, but the reviewers were not sure what the necessary experiments would be. To gauge twin-spot clone sizes? Please address.

We agree that clone fitness is an important consideration for MAGIC experiments. To test it, we generated WT clones in the wing imaginal disc using nMAGIC and quantified the sizes of the twin spots (BFP/BFP and WT/WT clones). Our results show that there is no statistical difference between these two types of clones. Thus, there is no intrinsic growth disadvantage to either type of mitotic clones generated by MAGIC.

(4) Include a schematic of the MAGIC method as Figure 1 or add it to Figure 1. Many may not be familiar with the method, so to promote its adoption, the authors should clearly introduce the MAGIC method in this paper (and not rely on readers to go to previous publications). For this paper to become a MAGIC reference paper, it should be self-contained.

We thank the reviewers for this suggestion. We have added diagrams that explain the principle of MAGIC and the difference between pMAGIC and nMAGIC in Figure 1 - Figure Supplement 1.

(5) Determine the utility of using a hs-Cas9 line for temporal induction of MAGIC clones. This is a traditional method for mitotic clone induction (with hsFLP/FRTs), and its use with the MAGIC system (especially pMAGIC) could also make it more attractive, especially to label small populations of neurons born at known times. To this point, the authors could generate pMAGIC clones using hs-Cas9 for commonly used adult target neurons, such as projection neurons, central complex neurons, or mushroom body neurons. The method to label small numbers of these adult neurons is well worked out with known GAL4 lines, and demonstrating that pMAGIC could have similar results would capture the attention of many not familiar with the pMAGIC method.

We agree that demonstrating inducible clone induction in the adult brain is an effective way for people to compare MAGIC with the MARCM method they are probably more familiar with. We used a heat shock Cas9 developed by the Tzumin Lee group (GarciaMarques, Espinosa-Medina et al. 2020) to experiment with clone induction, and the results are shown in the new Figure 3 (J-K’). We show that, with a pan-neuronal Gal4, heat shock during wandering 3rd instar larval stage induced more clones than during the pupal stage, and the later heat shock readily produced sparsely labeled neurons whose single-cell morphology can be easily visualized.

Reviewer #1 (Recommendations for the authors):

This is a marked improvement over the existing methods that the authors' lab has previously generated. It will be a nice addition to the Drosophila genetic tool kit after minor revisions.

We appreciate the reviewer’s recognition of the new tools we developed.

Minor issues:

(1) In the data in Figures 1G and H, it is not ideal to compare the effect of different modifications on two different transgenes. uH and uDEH are compared in gRNA-40D2, whereas uDEH, tDEH, and tDES are compared in gRNA-42A4. If the transgenics are already available, it would be better to compare the uH, uDEH, tDEH, and tDES on either gRNA-40D2 or gRNA-42A4.

We appreciate the reviewer’s concern. These transgenes were developed during different phases of this project. We first adopted the uDEH design during improvement of gRNA40D2, which solved both the leaky activity of pxn-Gal4 and dim epidermal clones. However, when we tried to expand this design to 2R (such as 42A4), we found that the clones were still too dim (probably due to positional effects). Thus, we next used uDEH in gRNA-42A4 as a base for further improvements. We did not make a uH version for gRNA-42A4 because we already knew that it is inferior to uDEH. Because of this history, we did not have the full set for gRNA42A4.

Despite the lack of uH for gRNA-42A4, we believe our comparisons of different designs are still valid, given that uH and uDEH were compared with identical sequences elsewhere in the transgenic vector (including the gRNA target sequence) and in the identical insertion site.

(2) It is not clear whether the authors tested destabilized Gal80 is potent to suppress Gal4 (e.g., in suppressing pxn>CD4-tdTom in hemocytes). The results in Figure 1C-E should be repeated with tub-Gal80-DE-SV40.

We apologize for omitting the transgene identities in these experiments. We have redone the experiment using gRNA-42A4-tDES and updated the figures to clearly indicate which transgenes were used.

(3) The difference in sgRNA scaffolds can be better explained in the text. The explanation here is very bare bones and reads like jargon. (i.e., changing F+E gRNA scaffold with gRNA2.1 scaffold is not a sufficient explanation).

We have added more explanations to the differences between the scaffolds as suggested.

(4) The stocks should be sent to Bloomington Stock Center to ensure widespread adoption of the method. This includes the Cas9 lines that are generated and used.

It is our plan to freely share the reagents developed in this study with the community. Most of the fly lines are already available at Bloomington (https://bdsc.indiana.edu/stocks/misc/magic.html and https://bdsc.indiana.edu/stocks/genome_editing/crispr_cas9.html). We are in the process of depositing the remaining ones to BDSC.

In conclusion, this is a nicely written manuscript that improves currently available tools and should be of interest to the readership of this journal.

Reviewer #2 (Recommendations for the authors):

Typos spotted:

Line 163 issues -> tissues

Line 613 significance -> significant

We thank the reviewer for catching these typos. We have corrected them.

Reviewer #3 (Recommendations for the authors):

This is a welcome update to the MAGIC system, which is a brilliant method that has not been as widely adopted as it should be. The authors validate and introduce updates to this system to increase clonal efficiency and more robust labeling (for both pMAGIC and nMAGIC). The data presented are robust and convincing.

We appreciate the reviewer’s positive comments about our study.

Suggestions to improve the presentation and adoption of this work:

(1) The MAGIC system might not be well known, and the manuscript would have benefited from an introductory schematic of how the system works. I realize this was already done in the PLoS Biology paper, but the authors should not assume readers will know that paper, or be willing to look it up. So a standalone schematic, as Figure 1, or something added to Figure 1, would greatly aid in understanding how this system works and what the new updates are doing.

We thank the reviewer for this kind suggestion. We have added diagrams that explain the principle of MAGIC and the difference between pMAGIC and nMAGIC in Figure 1 - figure supplement 1.

(2) There were many instances where abbreviations were not clearly defined, especially in the Figures and Figure legends. The main text is well-written, and while the information is in there, it is beneficial when the Figures and Figure legends can stand alone. For example:

(a) Figure 1. DE, not defined in the Figure or Figure legend.

(b) Figure 1. 'p' and 'n' not defined in the Figure legend.

(c) The different Cas9 lines or GAL4 lines used-a brief description of their expression patterns might be helpful in the legend. E.g., zk-Cas9, vas-Cas9, gcm-Cas9, R38F11-GAL4, RabX4Gal4.

We apologize for omitting the details mentioned. They have been added to the figures and figure legends.

(3) "Traditional" mosaic analyses took advantage of hsFLP for inducible induction and to control the number of mitotic clones that were induced. A hs-Cas9 line does exist (as correctly pointed out by the authors), and it would be a valuable addition if the authors tested the utility of this reagent with the MAGIC system. Many possible adopters may not like the idea that an alwayson Cas9 line is used, which could result in too many clones, especially if one wanted to label very few cells. Granted, one could use a 'worse' gRNA-Marker line as mentioned in the manuscript, but this might still be hard to titrate, as well as an inducible system that uses a heatshock promoter. A hs promoter is especially useful for birthdating cells during development.

We thank the reviewer for suggesting this experiment. We agree that demonstrating inducible clone induction in the adult brain is an effective way for people to compare MAGIC with the MARCM method they are probably more familiar with. We used a heat shock Cas9 developed by the Tzumin Lee group (Chen et al., Development, 2020) to experiment with clone induction, and the results are shown in the new Figure 3 (K and J). We show that, with a panneuronal Gal4, heat shock during wandering 3rd instar larval stage induced more clones than during the pupal stage, and the later heat shock readily produced sparsely labeled neurons whose single-cell morphology can be easily visualized.

(4) Lines 61-63. "However, most of these mutant chromosomes cannot be analyzed by traditional mosaic techniques due to the lack of FRT sites or incompatibility with the FRT/Flp system." It might also be worth mentioning that recombining existing reagents (e.g., mutants, etc) onto an FRT chromosome can be labor and time-intensive. A brilliant advantage of MAGIC is that it can be used with any existing stock, such as from classical EMS mutant screens, Df screens (as pointed out), etc. So the more the authors can emphasize a new way of thinking (e.g, you don't need to recombine your mutant of interest onto an FRT stock before you can get started), the better!

We thank the reviewer for this kind suggestion. As suggested, we have expanded our introduction and discussion to emphasize the advantages of the MAGIC system over traditional mosaic techniques.

(5) One incredible advantage of the MAGIC system is that it can direct where recombination occurs. So if one had two mutations on a chromosome arm, it could be possible to make the most distal homozygous mutant while the other remains heterozygous. This is not possible with current FRT-based methods. It's not necessary to demonstrate this, but perhaps the authors could mention it as a possible next step? This was somewhat implied by lines 66-67 "In comparison, MAGIC can potentially be used to study these genes because the crossover site in MAGIC can be flexibly defined by users".

Again, we thank the reviewer for this nice suggestion. We have added this point to the discussion.

(6) How stable are the MAGIC lines? If gRNA (with Cas9 expressed) induced a germline mutation of the target site, the MAGIC line would break down. How often is this observed? Some mention of this would be appreciated, especially to end users, if caution is necessary and gRNA-marker stocks should not be maintained in the same flies as an x-Cas9 line.

The reviewer made a very important point. Keeping gRNA and Cas9 in the same strain will risk mutating the target sequence in the germline, if the Cas9 has any activity in the germline. Thus, it is not recommended to keep gRNA and Cas9 in the same flies over multiple generations. For MAGIC experiments, this concern is lessened because by crossing gRNA + Cas9 flies to another strain containing the chromosome of interest, clones can still be induced (possibly with less efficiency) because the chromosome of interest is still cuttable by Cas9. Nevertheless, to address this concern, we have recently developed anti-CRISPR tools to suppress Cas9 activity in such strains. These tools will be reported in a separate study.

In the revised manuscript, we added this point in Discussion to caution users.

(7) Line 157, "identify efficient gRNAs for every chromosomal arm.". What is considered "efficient"? Is this quantifiable? Eg., >= 10 clones.

Thanks for pointing this out! “Efficient” is an arbitrary evaluation, as different experiments may require different efficiencies. But operationally, we consider any gRNA that can generate >= 10 neuronal clones per larva as being efficient. We have clarified it in the text.

(8) Line 163, "highly packed _issues_ such as the brain"; spelling, should be "tissues"

Thanks for catching this typo. It has been corrected.

(9) The authors use ey-Cas9 for their demonstration of adult brain labeling. Additional adult brain examples would increase exposure of this method and attract wider attention- targeting structures that have been well characterized, such as projection neurons (GH146-GAL4), central complex, mushroom bodies, etc. Especially if hs-Cas9 could be utilized to mimic previous MARCM clones (for example).

We thank the reviewer for suggesting heat shock-induced clones in the adult brain. We have conducted the experiment as explained above and shown in Figure 3J-3K’. We showed a single neuronal clone that resembles lateral horn Leucokinin neurons.

(10) Line 216, "Despite these advances, existing mutations on FRT-lacking 4th chromosomes still cannot be analyzed by the FRT/Flp system." For context, it might be worth pointing out that meiotic recombination is exceedingly rare on the 4th chromosome, which means it is practically impossible to recombine existing 4th chromosome mutations onto an FRT chromosome.

We thank the reviewer for this kind suggestion. We have added a note about the difficulty of recombining FRT onto the 4th chromosome.

(11) Figure 2 legend. What is the full genotype for D and E? eg, what is RabX4>MApHS?

We apologize for being brief with the details. RabX4-Gal4 is a pan-neuronal driver. UAS-MApHS is a membrane fluorescent marker (UAS-pHluorin-CD4-tdTom). The genotypes have been added to the figure legend.

(12) It would be good to include the Bloomington Stock numbers for the MAGIC toolkit, especially in Table 1. And include an HTML reference to their MAGIC page at Bloomington

(https://bdsc.indiana.edu/stocks/misc/magic.html).

Thank you for this suggestion! We have done as suggested.

(13) Similarly, the key plasmids to create the improved gRNA-marker insertions should be deposited to Addgene (or similar repository) and their ID numbers included in the resources table.

The plasmids have been deposited to Addgene and are currently being validated.

(14) The authors might consider including (perhaps as supplementary to Figure 1 or Figure 2) a crossing scheme for one of their MAGIC experiments. This will make it even clearer how a MAGIC experiment could be set up using existing fly reagents.

This is a good suggestion! We have added an example crossing scheme in Figure 1 – figure supplement 1C.

Version 2

Abstract

Mosaic analysis has been instrumental in advancing developmental and cell biology. Most current mosaic techniques rely on exogenous site-specific recombination sequences that need to be introduced into the genome, limiting their application. Mosaic analysis by gRNA-induced crossing-over (MAGIC) was recently developed in Drosophila to eliminate this requirement by inducing somatic recombination through CRISPR/Cas9-generated DNA double-strand breaks. However, MAGIC has not been widely adopted because gRNA-markers, a required component for this technique, are not yet available for most chromosomes. Here, we present a complete, genome-wide gRNA-marker kit that incorporates optimized designs for enhanced clone induction and more effective clone labeling in both positive MAGIC (pMAGIC) and negative MAGIC (nMAGIC). With this kit, we demonstrate clonal analysis in a broad range of Drosophila tissues, including cell types that have been difficult to analyze using recombinase-based systems. Notably, MAGIC enables clonal analysis of pericentromeric genes and deficiency chromosomes and in interspecific hybrid animals, opening new avenues for gene function study, rapid gene discovery, and understanding cellular basis of speciation. This MAGIC kit complements existing systems and makes mosaic analysis accessible to address a wider range of biological questions.

Introduction

Mosaic animals containing genetically distinct populations of cells in the same organism are useful for in vivo studies of complex biological processes. For this reason, techniques that can generate genetically labeled mosaic clones have been utilized in both vertebrates and invertebrates to study tissue-specific functions of pleiotropic genes, developmental timing, cell lineages, cell proliferation, neural wiring, and many other biological phenomena (Xu and Rubin 2012, Griffin, Binari et al. 2014, Germani, Bergantinos et al. 2018). The most popular mosaic techniques rely on site-specific recombination systems, such as FRT/Flp (Golic and Lindquist 1989, Xu and Rubin 1993) and LoxP/Cre (Zong, Espinosa et al. 2005, Henner, Ventura et al. 2013), to induce somatic recombination between homologous chromosomes. Such techniques require the introduction of recombination sites to specific locations in the genome and thus cannot be applied to unmodified chromosomes. To overcome this limitation, we recently developed a recombinase-independent mosaic technique called mosaic analysis by gRNA-induced crossing-over (MAGIC) (Allen, Koreman et al. 2021). In MAGIC, the CRISPR/Cas9 system generates double-strand breaks (DSBs) at a predefined genome location to induce homologous recombination in precursor cells during S/G2 phase. Subsequent chromosomal segregation during mitosis can result in clones homozygous for the chromosomal segments distal to the crossover site (Figure 1 - figure supplement 1A). Two variants of this technique in Drosophila, positive MAGIC (pMAGIC) and negative MAGIC (nMAGIC), label the resulting homozygous clones by the presence and absence of fluorescent markers, respectively (Figure 1 - figure supplement 1B) (Allen, Koreman et al. 2021). Like FRT/Flp-based techniques, MAGIC enables characterization of homozygous clones of lethal mutations in otherwise heterozygous animals (Allen, Koreman et al. 2021) (e.g. Figure 1 - figure supplement 1C).

New gRNA-marker designs improve pMAGIC and nMAGIC.

(A) Original and new designs of gRNA-markers for pMAGIC and nMAGIC. (B) Comparison of clone frequency in larval sensory neurons between two gRNA designs. Clones were induced by zk-Cas9 (expressed in the embryonic ectoderm) and labeled by the pan-neuronal driver RabX4-Gal4>MApHS (MApHS: pHluorin-CD4-tdTomato). The number represents clones between A1 and A7 segments on one side of each larva. n = larvae number: tgFE (n=10), Qtg2.1 (n=10). (C-E) Labeling of hemocytes in whole 3rd instar larvae by pxn-Gal4>CD4-tdTom alone (C) or together with gRNA-42A4(Gal80)-uDEH (ubi-Gal80) (D) or gRNA-42A4(Gal80)-tDES (tub-Gal80) (E). The panels on the right show enlarged views of the boxed regions. (F) Designs of Gal80 variants tested in pMAGIC gRNA-markers. (G) The brightness of epidermal clones induced by zk-Cas9 and labeled by the epidermal driver R38F11-Gal4>tdTom in the presence of pMAGIC gRNA-markers. n = image numbers: gRNA-40D2-uH (n = 32), gRNA-40D2-uDEH (n = 31), gRNA-42A4-uDEH (n = 52), gRNA-42A4-tDEH (n = 39), gRNA-42A4-tDES (n = 38). (H) The brightness of neuronal clones induced by zk-Cas9 and labeled by RabX4-Gal4>MApHS in the presence of pMAGIC gRNA-markers. The brightness of tdTom was measured and compared. n = neuron numbers: gRNA-40D2-uH (n = 16), gRNA-40D2-uDEH (n = 16), gRNA-42A4-uDEH (n = 16), gRNA-42A4-tDEH (n = 15), gRNA-42A4-tDES (n = 16). (I) A portion of a larval wing disc containing nMAGIC clones visualized by nlsBFP. (J and J’) A portion of a wing disc containing nMAGIC clones labeled by cytosolic BFP (J) and HA staining (J’). (K) Epidermal clones on the larva body wall labeled by nlsBFP. (L) Epidermal clones visualized by cytosolic BFP. (M) A portion of a wing disc containing nMAGIC clones labeled by cytosolic miRFP680 (IFP). (O) Sizes of nMAGIC BFP/BFP clones and wildtype (+/+) clones in wing discs. Two types of clones in the same discs were connected. n = wing disc number: BFP/BFP (n = 18), +/+ (n=18). In all plots, black bar, mean; red bar, SD; AU, arbitrary unit. Student’s t-test in (B); one-way analysis of variance (ANOVA) and Tukey’s honest significant difference (HSD) test in (G) and (H). paired t-test in (O) *p≤0.05, **p≤0.01, ***p≤0.001, ns, not significant. For (C-E), scale bar, 300 µm. For (I-M), scale bar, 100 µm.

However, unlike FRT/Flp-based techniques, MAGIC does not require prior genetic modification of the test chromosome. Thus, it can potentially be used on any chromosome and have much wider applications. Foremost, mutations of diverse natures have been established for most Drosophila genes (Hacker, Nystedt et al. 2003, Bellen, Levis et al. 2004, Thibault, Singer et al. 2004, Staudt, Molitor et al. 2005, Bellen, Levis et al. 2011, Yamamoto, Jaiswal et al. 2014), and thousands of deficiency strains harbor deletions that collectively uncover 98.4% of the Drosophila genome (Cook, Christensen et al. 2012). However, most of these mutant chromosomes cannot be analyzed by traditional mosaic techniques due to the lack of FRT sites or incompatibility with the FRT/Flp system. Although FRT sites can be introduced onto mutant chromosomes through genetic recombination, this process is labor-intensive and time-consuming and thus is impractical at a large scale. In contrast, MAGIC can theoretically be applied to any existing stock, including those from classical mutagenesis screens and deficiency libraries, allowing convenient genome-wide mosaic screens. In addition, genes located more proximal to centromeres than existing FRT sites cannot be analyzed by FRT/Flp techniques. In comparison, MAGIC can potentially be used to study these genes because the crossover site in MAGIC can be flexibly defined by users. Lastly, given that MAGIC is compatible with wild-derived chromosomes (Allen, Koreman et al. 2021), it may be able to generate homozygous clones of a chromosome derived from a single species in an interspecific hybrid animal, allowing the study of species-related cell-cell interactions.

Despite these potentials, MAGIC has not been widely adopted by the Drosophila community. A major barrier is the lack of gRNA-marker transgenes on most chromosomal arms, which are necessary for DSB induction and fluorescent clone labeling (Allen, Koreman et al. 2021). In addition, the existing gRNA-markers suffer from several limitations, including low frequency of clone induction, weak labeling of pMAGIC clones, and suboptimal visualization of nMAGIC clones. Because of these reasons, MAGIC has only been successfully applied to a few genes in a limited number of Drosophila tissues (Allen, Koreman et al. 2021, Chen, Wang et al. 2025).

To overcome these limitations, here we first optimized gRNA-marker designs to improve clonal induction, the brightness of pMAGIC clones, and visualization of nMAGIC clones. Then we generated pMAGIC and nMAGIC gRNA-markers for all chromosomal arms and characterized their ability to generate clones. Using this kit, we demonstrate mosaic analysis of centromere-proximal genes, deficiency chromosomes, chromosomes derived from different Drosophila species in interspecific hybrid animals. This kit allows optimal clone induction in diverse cell and tissue types and should be useful for studying a wide range of biological processes.

Results

New gRNA-marker designs improve pMAGIC and nMAGIC

MAGIC relies on a gRNA-marker dual-transgene (Figure 1A and Figure 1 - figure supplement 1B) inserted in a specific chromosomal arm to both induce and visualize clones homozygous for this arm (Allen, Koreman et al. 2021). The gRNA part of the construct expresses two gRNAs ubiquitously to target a pericentromeric location on this arm. Two gRNAs targeting two sequences that are close to each other, instead of a single gRNA, are used to increase the probability of DSBs and thus the clone frequency. The marker part in pMAGIC utilizes a ubiquitously expressed Gal80 to prevent Gal4-dependent labeling of heterozygous and homozygous cells for the gRNA-marker, while allowing labeling of homozygous cells that lose the gRNA-marker. In contrast, nMAGIC uses ubiquitously expressed BFP, which results in brighter labeling of gRNA-marker homozygous cells, intermediate labeling of gRNA-marker heterozygous cells, and lack of labeling of gRNA-marker-negative homozygous cells (Figure 1 - figure supplement 1B).

We have previously developed gRNA-marker vectors for both pMAGIC and nMAGIC. However, gRNA-markers made with these vectors exhibit some limitations. First, the clone frequency can be low for certain gRNAs (Allen, Koreman et al. 2021). Second, the ubi-Gal80 in pMAGIC gRNA-markers does not completely suppress Gal4 activity in certain tissues, such as hemocytes. Third, pMAGIC clones are sometimes too dim to visualize cell morphology, such as thin dendrite and axon projections of neurons. Lastly, nMAGIC utilizes a nuclear BFP marker (nlsBFP), which does not show the cell shape and sometimes cannot mark clones effectively. Thus, to optimize MAGIC, we first sought to improve the gRNA-marker designs.

Since clone induction in MAGIC depends on gRNA-induced DNA DSBs, we asked whether a more efficient gRNA design enhances clone frequency in somatic tissues. The previous gRNA-marker vectors used a tgFE design, which contains a flip of A-U positions and a stem-loop extension (F+E) of the original gRNA scaffold and a tRNAGly before each gRNA targeting sequence (Figure 1A) (Allen, Koreman et al. 2021). However, we have shown that an improved Qtg2.1 design, which contains an additional extension of the second stem loop in the gRNA scaffold (gRNA2.1) and a single tRNAGln spacer between the two gRNAs (Figure 1A), is much more mutagenic than tgFE in somatic tissues (Koreman, Xu et al. 2021). We thus compared two pMAGIC gRNA-markers that are based on these two gRNA designs but target the same genomic sequences at cytological band 40D2 to assess their ability to induce clones in peripheral sensory neurons. When used with the same neuronal/epidermal precursor Cas9, zk-Cas9 (Allen, Koreman et al. 2021), Qtg2.1 resulted in three times more neuronal clones as compared to tgFE (Figure 1B), confirming a positive correlation between gRNA efficiency and clone frequency.

To ensure more complete suppression of Gal4 activity by Gal80 in pMAGIC, we replaced the ubi enhancer driving Gal80 expression with an αTub84B (tub) enhancer that was used in tub-Gal80 in the MARCM system (Lee and Luo 1999). When combined with the larval hemocyte marker pxn-Gal4 UAS-CD4-tdTom (Han, Song et al. 2014) (Figure 1C), ubi-Gal80 did not completely suppress the labeling of hemocytes (Figure 1D). In contrast, no labeled hemocytes could be detected in the presence of tub-Gal80 (Figure 1E), suggesting that tub-Gal80 is a better marker for pMAGIC.

To improve clone brightness in pMAGIC, we sought to destabilize Gal80 and reduce its expression, reasoning that dim clones are due to prolonged Gal80 activity after clone induction. In addition to replacing the ubi enhancer with the tub enhancer, we also introduced protein and mRNA destabilization sequences (DE) (Zubiaga, Belasco et al. 1995, Li, Zhao et al. 1998) at the 3’ end of the Gal80 coding sequence, and replaced the His2Av polyA by SV40 polyA (Figure 1F), as the latter reduces transgene expression (Han, Jan et al. 2011). By measuring the brightness of epidermal (Figure 1G) and neuronal (Figure 1H) clones induced by gRNA-40D2 and gRNA-42A4, we found that each of the three changes improved clone brightness.

Finally, to visualize cell shape in nMAGIC, we replaced nlsBFP with cytosolic BFP tagged by 3X Hemagglutinin (HA), in addition to utilizing the tub enhancer. This new design allowed us to better discern clone shapes in both the wing imaginal disc and the epidermis (Figure 1J, 1J’, and 1L), in contrast to the previous ubi-nlsBFP design (Figure 1I and 1K). To make nMAGIC compatible with more fluorescent reagents, we generated an additional vector that contains miRFP680, a far-red/infrared fluorescent protein (IFP) (Matlashov, Shcherbakova et al. 2020), in place of BFP. Wing-disc clones labeled with this marker were readily detectable in unstained tissues (Figure 1M). By measuring the sizes of homozygous gRNA-marker clones (BFP/BFP) and homozygous wildtype (WT) clones (+/+) in wing discs, we found that these two cell populations in the twin spots showed no noticeable bias in growth or viability (Figure 1O).

Thus, by altering the designs of the gRNAs and the Gal80 and BFP markers, we created new vectors optimized for more robust applications of nMAGIC and pMAGIC.

A gRNA-marker kit is established for all four chromosomes of Drosophila

To enhance the utility of MAGIC in Drosophila, we generated complete sets of pMAGIC and nMAGIC (BFP version) gRNA-markers for all four chromosomes (Figure 2A). To identify suitable gRNA target sites, we analyzed the pericentromeric sequences of X, 2L, 2R, 3L, 3R, and 4, based on three criteria (Allen, Koreman et al. 2021): conserved in closely related Drosophila species to minimize the chance of single nucleotide polymorphism, (2) located away from functionally critical regions to avoid disrupting essential processes, and (3) unique within the genome to minimize off-target effects. For each MAGIC construct, we selected a pair of non-repetitive gRNA target sequences in intergenic regions to maximize the chances of DSBs. These two sequences are closely linked to minimize the risk of large deletions. Given the variable efficiency of gRNA target sequences, we selected three pairs of gRNAs targeting three chromosomal locations for each chromosomal arm and named them according to the corresponding cytoband (Table 1).

A genome-wide gRNA-marker kit suits diverse needs of clone frequency.

(A) Scheme of gRNA-marker insertion sites and target sites on Drosophila chromosomes. (B) Comparison of clone frequencies of all pMAGIC gRNA-markers in larval sensory neurons, clones are labeled using RabX4-Gal4>MApHs (for Chromosome X, II and IV) or 21-7-Gal4 UAS-MApHS (for Chromosome III). n = larvae number: X2 (n = 10), 20F2 (n = 10), 20F1(n = 10), 40D2 (n = 20), 40D4 (n = 10), 40E1 (n = 10), 41F9 (n = 20), 41F11 (n = 10), 42A4 (n = 10), 80C1 (n = 20), 80C2 (n = 14), 80F5 (n = 15), 81F (n = 10), 82A4 (n = 10), 82C3 (n = 10), 101F1a (n = 10), 101F1b (n = 10), 101F1c (n = 10). (C) Comparison of clone areas in larval wing discs labeled by nMAGIC gRNA-markers on 2R. n = wing disc number: 41F9 (n = 14), 41F11 (n = 16), 42A4 (n = 15). (D and E) Neuronal clones in the central part of the adult brain induced by ey-Cas9 (expressed in progenitor cells of many neuronal tissues) and labeled by RabX4-Gal4>MApHS along with pMAGIC gRNA-markers gRNA-40D2 (D) and gRNA-40E1 (E). MApHS contains pHluorin and tdTom (Han, Song et al. 2014), but only the tdTom channel is shown. In all plots, black bar, mean; red bar, SD. One-way ANOVA and Tukey’s HSD test. *p≤0.05, **p≤0.01, ***p≤0.001, ns, not significant. For (D) and (E), scale bar 100 µm.

To evaluate the clone-induction properties of these gRNAs, we combined the pMAGIC set with zk-Cas9 and counted the number of neuronal clones in A1-A7 larval hemi-segments (Figure 2B). As expected, the clone frequency varied from gRNA to gRNA, but we were able to identify efficient gRNAs (>= 10 clones per larva) for every chromosomal arm. We previously found that different gRNAs follow the same trend of relative efficiency in different tissues (Allen, Koreman et al. 2021). Here we additionally tested nMAGIC gRNAs for 2R in the wing disc (Figure 2C) and noticed a similar trend of clone induction to that of their pMAGIC counterparts in sensory neurons (Figure 2B), suggesting that the results in sensory neurons are transferrable to other tissues.

Certain gRNAs (e.g. gRNA-40E1) exhibited very low clone frequency. Such gRNAs can be useful for inducing sparse clones in highly packed tissues such as the brain. For example, using the same ey-Cas9 (Ji, Sapar et al. 2022), the highly efficient gRNA-40D2 induced too many neuronal clones in the adult brain for morphological analysis (Figure 2D), while gRNA-40E1 gave rise to few clones, whose projection patterns were much easier to analyze (Figure 2E).

Together, the pMAGIC and nMAGIC gRNA-marker lines constitute a complete kit (Table 1) for genome-wide MAGIC applications in Drosophila.

MAGIC allows clonal analysis in diverse tissues and cell types

To determine if MAGIC can be applied to diverse tissues in Drosophila, we conducted clonal analysis in the larva using gRNA-40D2(Gal80) and several tissue-specific Cas9s. With the ubiquitous vas-cas9 (Lopez Del Amo, Juste et al. 2022) and tub-Gal4>UAS-mCD8-GFP, we readily detected clones in the larval brain (Figure 3A), proliferating tissues like eye and leg discs (Figure 3B-C), and polyploid tissues like the fat body, gut, and trachea (Figure 3D-F). Clone induction in polyploid tissues suggests that crossing-over events occurred before the last cell division. Using zk-cas9 and R38F11-Gal4>UAS-tdTomato (tdTom), we observed frequent epidermal clones (Figure 3G). Using the glia precursor gcm-Cas9 and repo-Gal4>UAS-mCD8-GFP, we detected individual glial clones in the brain (Figure 3H). The new pMAGIC gRNA-marker design allowed us to reliably induce hemocyte clones (Figure 3I).

MAGIC allows clonal analysis in diverse tissues and cell types.

(A-F) pMAGIC clones induced in different tissues by vas-Cas9 (ubiquitous Cas9) gRNA-40D2(Gal80) and labeled by tub-Gal4 UAS-CD8-GFP (green). DAPI staining (white) shows all nuclei. (G) A pMAGIC epidermal clone on the larval body wall induced by zk-Cas9 gRNA-40D2(Gal80) and labeled by R38F11-Gal4>tdTom (green). Epidermal junctions are labeled by α-Catenin-GFP (white). (H) pMAGIC glia clones in the larval brain induced by gcm-Cas9 (expressed in glial precursor genes) gRNA-40D2(Gal80) and labeled by repo-Gal4 UAS-CD8-GFP (green). Glial nuclei are labeled by Repo staining (white). (I) pMAGIC hemocyte clones induced by Act-Cas9 gRNA-40D2(Gal80) and labeled by pxn-Gal4>CD4-tdTom. (J-K’) pMAGIC clones in adult brain induced by hs-Cas9 gRNA-40D2(Gal80) and labeled by RabX4-Gal4>MApHS. Heat shock was performed at 120 h after egg lay (AEL) (J-J’) and 48 h after puparium formation (APF) (K-K’). The boxed areas were enlarged to show clones in the mushroom body and lateral horn region. Only the tdTom channel is shown. In (A), (D-F), (H), (J), and (K), scale bar 100 µm; in (B-C), (G), (J’), and (K’), scale bar 50 µm; in (I), scale bar 25 µm.

The ability to control the timing of clone induction has been instrumental for neuronal birth dating and modulation of clone frequency in traditional MARCM analysis of the Drosophila adult brain. To explore the potential of pMAGIC to serve similar purposes, we used heat-shock (hs) Cas9 (Garcia-Marques, Espinosa-Medina et al. 2020), along with gRNA-40D(Gal80) and a pan-neuronal Gal4/UAS-membrane marker combination (RabX4-Gal4>UAS-MApHS), to induce neuronal clones in the adult brain. Heat shock at 120 h after egg laying (AEL) induced too many clones for separating individual neurons (Figure 3J-J’), while later heat shock at 48 h after pupal formation (APF) produced many fewer, spatially separated clones in the central brain. These results collectively demonstrate MAGIC’s efficacy and flexibility in generating clones in diverse Drosophila tissues, indicating its value in studying gene function and cell lineage in various tissue contexts.

The neuromuscular junction (NMJ) in Drosophila larvae has been a valuable model for elucidating synaptic biology, owing to its simplicity, accessibility, and conserved features shared with mammalian excitatory synapses (Menon, Carrillo et al. 2013, Charng, Yamamoto et al. 2014, Chou, Johnson et al. 2020). Despite its popularity, the NMJ has rarely been analyzed by the MARCM technique in the literature, likely due to the difficulty of inducing clones in motor neurons. To test the effectiveness of MAGIC in analyzing gene function at the NMJ, we selected two genes crucial for synaptic function and construction: Vesicular glutamate transporter (VGlut) (Daniels, Collins et al. 2004) and bruchpilot (brp) (Kittel, Wichmann et al. 2006, Wagh, Rasse et al. 2006), null mutations of each of which exhibit recessive lethality in larvae. Combined with appropriate gRNA(Gal80) lines (42A4 for VGlut on 2R and 40D2 for brp on 2L) and tub-Gal4 UAS-CD8-GFP, zk-Cas9 induced frequent clones in type Ib boutons (Figure 4A-B”). The loss of VGlut or Brp specifically in GFP-labeled NMJs was confirmed by immunostaining. These results exemplify the value of pMAGIC for dissecting gene function in NMJ biology at the single cell level.

MAGIC facilitates clonal analysis at the NMJ.

(A-A”) pMAGIC clones of VGlut1 mutation in motor neurons at the neuromuscular junction. Clones were induced by zk-Cas9 gRNA-40D2(Gal80) and labeled by tub-Gal4>CD8-GFP. The loss of VGlut is confirmed by VGlut staining. The mutant clones are outlined in (A”). (B-B”) A pMAGIC clone of brpd09839 mutation in a motor neuron at the neuromuscular junction. Clones were induced by zk-Cas9 gRNA-42A4(Gal80) and labeled by tub-Gal4>CD8-GFP. The loss of Brp is confirmed by Brp staining. The mutant clone is outlined in (B”). In both experiments, HRP staining shows all axons. Scale bars, 10 µm.

MAGIC enables clonal analysis of pericentromeric genes, 4th chromosome-associated mutations, and in interspecific hybrid animals

Because all FRT sites in the existing FRT/Flp mosaic system are located at some distances away from centromeres, it has not been possible to study genes located between the FRT sites and the corresponding centromeres by clonal analysis. In contrast, the gRNA target site in a MAGIC experiment can be user-selected to enable clonal analysis of any given gene. To illustrate this potential, we used gRNA-41F9 to induce homozygous mutant clones of Ecdysone receptor (EcR), which is located at 42A10 and is inaccessible to all FRT sites on 2R. EcR is a transcription factor required for neuronal remodeling during metamorphosis (Brown, Cherbas et al. 2006). In peripheral sensory neurons, EcR activation at the beginning of pupariation causes apoptosis of some dendritic arborization (da) neurons while triggering dendrite pruning of other da neurons (Williams and Truman 2005). As expected, a wildtype (WT) pMAGIC clone of class IV da (C4da) neuron exhibited complete dendrite pruning at 16 h after puparium formation (APF), accompanied by dendrite debris phagocytosed into epidermal cells (Han, Song et al. 2014) (Figure 5A). In contrast, EcR mutant da neuron clones still maintained larval dendritic arbors at 16 h APF (Figures 5B-D), instead of dying (in the case of C3da) or undergoing dendrite pruning (in the cases of C1da and C4da).

MAGIC enables clonal analysis of pericentromeric genes, 4th chromosome-associated mutations, and in interspecific hybrid animals.

(A) A WT pMAGIC class IV da neuron clone exhibiting complete dendrite pruning at 16 hours APF. (B-D) pMAGIC clones of EcRM554fsmutation in da neurons imaged at 16 hours APF, exhibiting the lack of pruning (B and D) or apoptosis (C). In (A-D), the clones were induced by zk-cas9 with gRNA-41F9(Gal80) and labeled by RabX4-Gal4>MApHS. Neuronal cell bodies are indicated by arrows. Only the tdTom channel is shown. The signals in epidermal cells (A) were due to engulfment of pruned dendrites by epidermal cells (Han, Song et al. 2014). (E and F) WT (E) and Df(4)ED6380 (F) pMAGIC clones in C4da neurons induced by zk-cas9 gRNA-101Fc(Gal80) and labeled by RabX4-Gal4>MApHS. Only the tdTom channel is shown. (G) Normalized dendrite length of WT clones and deficiency clones. Black bar, mean; red bar, SD. Student’s t-test. ***p≤0.001. (H) Scheme for interspecific crosses between D. melanogaster (D.m) and D. simulans (D.s). (I and J) Wing discs from male (I) and female (J) progeny carrying clones. Scale bars, 50 µm.

Mosaic analysis of genes located on the 4th chromosome had not been possible until recent introduction of FRT sites onto this chromosome by CRISPR-mediated knock-in (Goldsmith, Shimell et al. 2022). Despite these advances, existing mutations on FRT-lacking 4th chromosomes still cannot be analyzed by the FRT/Flp system, given that meiotic recombination is exceedingly rare on the 4th chromosome, preventing introduction of FRT sites onto mutant chromosomes. A valuable gene-disruption resource in Drosophila is the deficiency library consisting of strains harboring chromosomal deletions. MAGIC can potentially be used in conjunction with the deficiency library to screen for genes important in a particular biological process. To test the potential of MAGIC for analyzing mutations on the 4th chromosome and for gene discovery with deficiencies, we generated pMAGIC C4da clones of Df(4)ED6380, a deficiency that deletes a segment between cytological bands 102B7 and 102D5. These clones show dramatically reduced dendrites compared to the WT controls (Figure 5E-5G), indicating the existence of genes important for dendrite growth in this region.

When examining nMAGIC gRNA-markers for the 4th chromosome, we noticed uneven expression of tub-3xHA-BFP in epidermal cells and some imaginal tissues, exemplified by cells lacking detectable BFP signal (Figure 5 – figure supplement 1A and B). This variegated expression is likely due to transgenes residing in heterochromatin and is common for transgenes located on the 4th chromosome (Riddle, Minoda et al. 2011). While this uneven expression limits the usefulness of our gRNA-markers in the corresponding epithelial tissues, Gal80 in pMAGIC gRNA-markers efficiently suppressed Gal4 activities in most neurons (Figure 5 – figure supplement 1C), conforming their effectiveness in neuronal MAGIC analysis.

Lastly, we wondered if the MAGIC system can generate clones in interspecific hybrid animals derived from D. melanogaster and D. simulans parents, given that the two species show a large degree of synteny (Chakraborty, Chang et al. 2021). The clones in these animals would contain homozygous chromosomal arms derived from a single species. To test this idea, we crossed D. melanogaster females containing gRNA-42A4(BFP); hh-cas9 to D. simulans males carrying a loss of function mutation of Lethal hybrid rescue mutation (Lhr), which ensures the viability of the hybrid male progeny (Barbash 2010). Indeed, we observed twin spots consisting of dark clones, containing only D. simulans 2R, and brighter clones, containing only D. melanogaster 2R, in wing discs of both female and male progeny (Figure 5G and 5H), demonstrating the feasibility of studying species-specific alleles in cell-cell interactions in interspecific hybrids.

Discussion

Conceptually, MAGIC is a simpler and more convenient mosaic technique compared to traditional recombinase-dependent methods. Theoretically, it can be applied directly to any existing stock, including those that are not currently compatible with the FRT/Flp system. However, the lack of gRNA-marker transgenes has prevented its wide application in Drosophila. In this study, we present a complete gRNA-marker kit that enables genome-wide MAGIC in Drosophila, removing this bottleneck. The new gRNA-markers incorporate optimized designs that improve clone frequency and labeling in both pMAGIC and nMAGIC. We further demonstrate the compatibility of MAGIC with broad tissues and cell types. More importantly, MAGIC enables mosaic analyses that could not be accomplished by existing FRT/Flp systems, such as those of pericentromeric genes, deficiency chromosomes, and interspecific homologous chromosomes. Thus, this MAGIC kit provides Drosophila researchers with greater flexibility for conducting mosaic analyses of diverse purposes.

To implement MAGIC, one needs to first choose a proper Cas9 that is expressed in the precursor cells of the targeted cell population. We show that ubiquitously expressed Cas9 lines, such as vas-Cas9 (Lopez Del Amo, Juste et al. 2022) and Act5C-Cas9 (Port, Chen et al. 2014), are sufficient to induce clones in broad tissues. Alternatively, heat shock (HS)-induced Cas9 (Garcia-Marques, Espinosa-Medina et al. 2020) can offer temporal control of clone induction in most tissues, akin to the HS-Flp commonly used in FRT-based mosaic analysis. The third option is a Cas9 driven by a tissue-specific enhancer specifically in the precursor cells of the target tissues. Examples shown in this study include ey-Cas9 expressed in many neuronal lineages (Ji, Sapar et al. 2022), gcm-Cas9 expressed in glial progenitor cells (Chen, Perry et al. 2024), zk-Cas9 expressed in precursor cells of epidermal cells, motor neurons, and somatosensory neurons (Koreman, Xu et al. 2021), and hh-Cas9 expressed in imaginal tissues (Poe, Wang et al. 2019). These Cas9 lines have the advantages of being more specific and efficient, and more convenient to use than ubiquitous or inducible ones. Multiple strategies, including enhancer fusion (Port, Chen et al. 2014, Poe, Wang et al. 2019), Gal4-to-Cas9 conversion (Chen, Yao et al. 2020, Koreman, Xu et al. 2021), and insertion of Cas9 in specific gene loci (Chen, Perry et al. 2024), have been developed to ease the generation of such tissue-specific Cas9 lines. As an ongoing effort, we have been converting Gal4 lines known to be expressed in progenitor cells into Cas9 (available at https://bdsc.indiana.edu/stocks/genome_editing/crispr_cas9.html), in the hope of making MAGIC accessible to more Drosophila tissues. It is worth noting that many Cas9 lines show leaky activity in the germline, which could mutate and inactivate the target sequence in the presence of a gRNA. Thus, it is not recommended to combine Cas9 and gRNA transgenes in the same strain as a long-term stock, unless a Cas9 inhibitor can also be introduced into this stock to suppress the germline activity.

The second component of MAGIC is a gRNA-marker line that resides on the appropriate chromosome arm and targets Cas9 to cut a pericentromeric site on the same arm. In the gRNA-marker kit, we generated three lines targeting different sites for each chromosome arm. These lines exhibit a broad range of clone frequencies in larval sensory neurons. As different gRNA-markers maintain similar relative efficiencies of clone induction across tissues (Allen, Koreman et al. 2021), one can choose a gRNA-marker more appropriate for their applications based on the target gene location and the desired clone frequency. Notably, a higher clone frequency may not always be beneficial, such as when studying the projection patterns of individual neurons in the brain. A low efficiency gRNA-marker in this kit may be more desirable in such applications.

Besides commonly conducted mosaic analysis in Drosophila, MAGIC also enables many novel analyses that are difficult or impossible to accomplish with traditional systems. One example is to analyze interactions among species-specific cells in an interspecific hybrid animal for understanding the cell biological basis of hybrid incompatibility. Interspecific crosses between WT D. melanogaster and WT D. simulans result in sex-specific lethality of F1 progeny (Barbash 2010). The cellular basis of this lethality is poorly understood. With MAGIC, one may generate mixed cell populations that are homozygous or heterozygous for species-specific alleles in the same hybrid embryo or larva, in which the relative fitness of the three cell populations can be compared. Similarly, MAGIC could complement genome-wide association studies (GWAS) using wild-derived isogenic strains and provide much more mechanistic detail. Such strains as those in the Drosophila Genetic Reference Panel (DGRP) (MacKay, Richards et al. 2012) have been very useful for discovering loci that are responsible for phenotypic variations. MAGIC can be combined with these strains to analyze the effects of homozygosity of specific variants at the cell biology level (Allen, Koreman et al. 2021). In addition, MAGIC can be combined with the Drosophila deficiency kit for rapid genome-wide mosaic screens. A deficiency exhibiting a desired phenotype in such screens can be further dissected with smaller deficiencies within the deleted region or existing mutations in candidate genes. In this way, the convenience of MAGIC could significantly accelerate phenotype-based gene discovery. Lastly, because the crossover site in MAGIC can be user-defined, one can easily generate gRNA-markers that induce somatic recombination at specific genome locations. With this feature, it is possible to induce crossover between two mutant alleles on the same chromosomal arm to generate clones homozygous for the distal mutation but heterozygous for the proximal mutation. This flexibility could also be useful for mapping undefined genetic loci that are responsible for certain phenotypes, especially in non-traditional model organisms.

Materials and methods

Fly Stocks and Husbandry

See the Key Resource Table (Supplementary File 1) for details of fly stocks used in this study. Most fly lines were either generated in the Han lab or obtained from the Bloomington Drosophila Stock Center. lethal hybrid rescue (lhr) mutant D. Simulans was a gift from Dr. Dan Barbash. All flies were grown on standard yeast-glucose medium, in a 12:12 light/dark cycle, at 25°C unless otherwise noted. Virgin males and females for mating experiments were aged for 3-5 days.

To generate and label pMAGIC clones in larval peripheral sensory neurons, we used either RabX4-Gal4 UAS-MApHS (for gRNA-markers on chromosomes X, II, and IV) or 21-7-Gal4 UAS-MApHS (for gRNA-markers on chromosome III) combined with zk-cas9. To count peripheral sensory neuronal clones on the larval body wall, third instar larvae were mounted laterally on slides and then counted in segment A1-A7 under a Nikon SMZ18 stereomicroscope. pMAGIC clones were induced and labeled by RabX4-Gal4 UAS-MApHS combined with ey-cas9 or hs-cas9 in the fly adult brain, by tub-Gal4 UAS-mCD8-GFP combined with vas-cas9 in larval brains, imaginal discs, fat bodies, guts, and trachea, by repo-Gal4 UAS-mCD8-GFP combined with gcm-cas9 in glia, by pxn-Gal4 UAS-tdTom combined with Act-cas9 in hemocytes, by tub-Gal4 UAS-mCD8-GFP combined with zk-cas9 in larval motor neurons, by R38F11-Gal4 UAS-tdTom combined with zk-cas9 in the larval epidermis. To induce nMAGIC clones in wing imaginal discs of interspecific hybrid animals, we use gRNA-42A4(BFP); hh-cas9 virgin females of D. melanogaster to cross with Lhr1 (Brideau, Flores et al. 2006) D. simulans males.

Molecular Cloning

MAGIC gRNA cloning vectors

pMAGIC gRNA-marker vectors constructed in this study include pAC-U63-gRNA2.1-ubiGal80(DE)-His2Av, pAC-U63-gRNA2.1-tubGal80(DE)-His2Av and pAC-U63-gRNA2.1-tubGal80(DE)-SV40. To make pAC-U63-gRNA2.1-ubiGal80(DE)-His2Av, a fragment containing gRNA2.1 scaffold (gRNA2.1) and U6 3’ flanking sequence (U63fl) was first used to replace gRNA2.1-QtRNA-Gal80(TS)-gRNA2.1-U63fl in pAC-U63-QtgRNA2.1-8R (Addgene 170514) (Koreman, Xu et al. 2021). Then a destabilization sequence (DE) containing a GS linker, amino acids (AAs) 422-461 of mouse ornithine decarboxylase (Zubiaga, Belasco et al. 1995), 2X RNA destabilizing nonamer (TTATTTATTgatccTTATTTATT) (Zubiaga, Belasco et al. 1995) was added to the C-terminus of Gal80 in frame. To make pAC-U63-gRNA2.1-tubGal80(DE)-His2Av, a 2.6 kb tub enhancer was amplified from pENTR221-tubP (Chen, Wang et al. 2025) by PCR and used to replace the ubi enhancer in pAC-U63-gRNA2.1-ubiGal80(DE)-His2Av. To make pAC-U63-gRNA2.1-tubGal80(DE)-SV40, a SV40 polyA sequence was amplified from pAPIC-PHCS (Han, Jan et al. 2011) and used to replace the His2Av polyA in pAC-U63-gRNA2.1-ubiGal80(DE)-His2Av. To make the nMAGIC gRNA-marker vector pAC-U63-gRNA2.1-tubBFP(HA), the gRNA2.1-U63fl fragment was first used to replace gRNA2.1-QtRNA-BFP(TS)-gRNA2.1-U63fl in pAC-U63-QtgRNA2.1-BR (Addgene 170513) (Koreman, Xu et al. 2021). The tub enhancer was then used to replace the ubi enhancer. Lastly, a synthetic DNA fragment (GenScript) encoding 3X HA was used to replace the nuclear localization signal at the N-terminus of mTagBFP. To make nMAGIC gRNA-marker vector pAC-U63-gRNA2.1-tub-miRFP680-T2A-HO1(HA), an miRFP680-T2A-HO1 coding sequence was used to replace BFP in pAC-U63-gRNA2.1-tubBFP(HA). HO1 encodes haeme oxygenase 1 and is necessary for generating the chromophore of miRFP680. Cloning was carried out by ligation with T4 ligase or NEBuilder DNA Assembly reactions (New England Biolabs Inc.).

MAGIC gRNA expression vectors

34 gRNA expression vectors were constructed with the corresponding gRNA cloning vectors as listed in Table 1 according to published protocols (Koreman, Xu et al. 2021). Briefly, for each expression vector, two primers containing appropriate gRNA target sequences (Supplementary File 2) were used to amplify a fragment consisting of 3’ end of U6:3 promotor, the first target sequence (TS1), gRNA2.1, tRNAQ, the second target sequence (TS2), and the beginning sequence of gRNA2.1 with pAC-U63-QtgRNA2.1-BR as the PCR template. The PCR product was then assembled with SapI-digested gRNA cloning vectors using NEBuilder DNA Assembly.

Injections were carried out by Rainbow Transgenic Flies (Camarillo, CA 93012 USA) or Genetivision (Stafford, TX 77477) to transform flies through φC31 integrase-mediated integration into attP docker sites. The 3xP3-RFP selection marker in gRNA-markers inserted at the attP[ZH-102D] site were removed by crossing to Cre.

Live Imaging

Live imaging of larval epidermal cells, sensory neurons, and hemocytes was performed as previously described (Poe, Tang et al. 2017). Animals were collected at 96 (for late third larvae) or 120 h (for wandering third instar larvae) AEL and mounted in glycerol on a slide with vacuum grease as a spacer. Animals were imaged using a Leica SP8 confocal microscope with a 40X NA1.3 oil objective, pinhole size 2 airy units, and a z-step size of 1 µm. For epidermis, images were taken at the dorsal midline of A2 and A3 segments. For dendritic arborization neurons, images were taken from A1 to A7 hemi-segments.

For imaging sensory neurons in pupae, newly formed pupae were collected and incubated at 25 °C. After 16 hours, the pupal cases were carefully removed, and the pupae were mounted dorsal side up on slides with halocarbon oil beneath a coverslip. Double-sided tape was used as a spacer. The mounted pupae were imaged using a Leica SP8 confocal microscope with a 40X NA1.3 oil objective.

Heat shock induction of neuronal clones in the adult brain

To induce heat shock, vials containing animals at the appropriate developmental stages were submerged in a 37°C water bath for 1 h. Subsequently, the vials were transferred back to a 25°C incubator until adults emerged.

Imaging of wing discs and other larval tissues

Larval dissections were performed as described previously (Poe, Wang et al. 2019). Briefly, wandering third instar larvae were dissected in a small petri dish filled with cold phosphate-buffered saline (PBS). The anterior half of the larva was inverted. To prepare imaginal discs, trachea and gut were removed. Samples were then transferred to 4% formaldehyde in PBS and fixed for 20 minutes at room temperature. After washing with PBS, the tissues were stained in DAPI (1:1000) in 0.2% PBST (PBS with 0.2% Triton X-100) for 5 minutes. The tissues were washed again in PBST and mounted in SlowFade Diamond Antifade Mountant (Thermo Fisher Scientific) on a glass slide. A coverslip was lightly pressed on top. Imaginal discs were imaged using a Leica SP8 confocal microscope with a 20X NA0.8 oil objective.

Adult brain imaging

Flies were aged for 1 day after eclosion. Brains were dissected in PBS at room temperature and then fixed in 4% paraformaldehyde in PBS with constant circular rotation for 20 minutes at room temperature. The brains were subsequently washed in 0.2% PBST and mounted on a glass slide under a glass coverslip. Vacuum grease was used as a spacer between the coverslip and the slide. Brains were imaged using a Leica SP8 confocal microscope with a 40X NA1.3 oil objective.

Larval fillet preparation

Larval fillet dissection was performed on a petri dish half-filled with PMDS gel. Wandering third instar larvae were pinned on the dish in PBS dorsal-side up and then dissected to expand the body wall. PBS was then removed and 4% formaldehyde in PBS was added to fix larvae for 15 minutes at room temperature. For VGlut staining, the fillets were fixed in Bouin’s solution for 5 minutes at room temperature. Fillets were rinsed and then washed at room temperature in PBS for 20 minutes or until the yellow color from Bouin’s solution faded. After immunostaining, the head and tail of fillets were removed, and the remaining fillets were placed in SlowFade Diamond Antifade Mountant on a glass slide. A coverslip was lightly pressed on top. Larval fillets were imaged using a Leica SP8 confocal microscope with a 40X NA1.3 oil objective.

Immunohistochemistry

Larval brains and larval fillets were rinsed and washed at room temperature in 0.2% PBST after fixation. The samples were then blocked in PBST with 5% normal donkey serum (NDS) for 1 hour before incubating with appropriate primary antibodies in the blocking solution. Brains were stained for 2 hours at room temperature, and fillets were stained overnight at 4°C. After additional rinsing and washing, the samples were incubated with secondary antibodies for 2 hours at room temperature. The samples were then rinsed and washed again before mounting and imaging. Primary antibodies used in this study are mouse anti-Repo antibody 8D12 (1:50 dilution), mouse anti-Brp antibody nc82 (1:100 dilution), rabbit anti-VGlut (1:200 dilution) (Chen, Perry et al. 2024), mouse anti-HA antibody (12CA5, 1:100), and goat anti-HRP conjugated with Cy3 (1:200). Secondary antibodies include donkey anti-mouse antibody conjugated with Cy5 (1:400) and donkey anti-rabbit antibody conjugated with Cy5 (1:400).

Image Analysis and Quantification

ImageJ analyses were conducted in Fiji/ImageJ. To compare brightness of clones in sensory neurons and epidermal cells, clones were detected based on thresholds to generate masks. The clone brightness within the masks was then measured as mean pixel intensity. To quantify clones in wing discs, 2-3 slices of optical sections in the middle of the disc were projected into a 2-dimensional (2D) image. Clones were detected based on a fixed threshold to generate masks, and the total area of clones in the wing pouch of each disc was measured.

The tracing and measurement of da neuron dendrites were done as previously described in detail (Poe, Tang et al. 2017). Briefly, dendrites were segmented using local thresholding. The segments were then converted into single-pixel-width skeletons. The total length of skeletons was calculated based on pixel distance. Normalized dendrite length was calculated as dendritic length (μm)/segment width (μm).

Statistical Analysis

One-way analysis of variance (ANOVA) with Tukey’s honest significant difference (HSD) test was used when dependent variable was normally distributed and there was approximately equal variance across groups. Student’s t-test or paired t-test was used when two groups are compared. For additional information on the number of samples, see figure legends. R studio was used for all statistical analyses.